---
title: "Data 624 Homework 9"
author: "Joao De Oliveira"
date: "5/3/2026"
output:
  pdf_document:
    toc: true
    toc_depth: 1
    highlight: tango
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 9,
  fig.height = 5.5,
  fig.align  = "center"
)
```


## Introduction
In this report, I will be answering questions 1, 2, 3 and 7 in the 8th chapter
from the Kuhn and Johnson book.


# Libraries

```{r libraries}
library(mlbench)
library(caret)
library(randomForest)
library(party)       
library(gbm)         
library(Cubist)      
library(rpart)       
library(rpart.plot)  
library(ggplot2)
library(tidyr)
library(gridExtra)
library(AppliedPredictiveModeling)  # ChemicalManufacturingProcess
```

---

# Question 8.1: Recreate the simulated data from Exercise 7.2


This exercise uses the same Friedman1 simulated data from Exercise 7.2.
The true equation is:

$$y = 10\sin(\pi x_1 x_2) + 20(x_3 - 0.5)^2 + 10x_4 + 5x_5 + \mathcal{N}(0,\sigma^2)$$

Only $x_1$ through $x_5$ are informative; $x_6$ through $x_{10}$ are pure noise.

## Simulate Data

```{r simulate-81}
set.seed(200)
simulated <- mlbench.friedman1(200, sd = 1)
simulated <- cbind(simulated$x, simulated$y)
simulated <- as.data.frame(simulated)
colnames(simulated)[ncol(simulated)] <- "y"
dim(simulated)
```

## (a) Fit a random forest model to all of the predictors, then estimate the variable importance scores:
> library(randomForest)
> library(caret)
> model1 <- randomForest(y ~ ., data = simulated,
+ importance = TRUE,
+ ntree = 1000)
> rfImp1 <- varImp(model1, scale = FALSE)

## Did the random forest model significantly use the uninformative predictors (V6 – V10)?

```{r rf-base, cache=TRUE}
set.seed(100)
model1 <- randomForest(y ~ ., data = simulated,
                       importance = TRUE,
                       ntree = 1000)

rfImp1 <- varImp(model1, scale = FALSE)
rfImp1_df <- data.frame(
  Predictor  = rownames(rfImp1),
  Importance = rfImp1$Overall,
  Informative = rownames(rfImp1) %in% paste0("V", 1:5)
)
rfImp1_df <- rfImp1_df[order(-rfImp1_df$Importance), ]
print(rfImp1_df)
```

```{r plot-rf-base}
ggplot(rfImp1_df,
       aes(x = reorder(Predictor, Importance),
           y = Importance, fill = Informative)) +
  geom_col(width = 0.6, color = "white") +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#4CAF50", "FALSE" = "#e07b54"),
                    labels = c("Non-informative (V6–V10)",
                               "Informative (V1–V5)")) +
  labs(title    = "Q8.1(a) – Random Forest Variable Importance",
       subtitle = "Uninformative predictors V6–V10 receive low but non-zero importance",
       x = "Predictor", y = "Importance (MSE increase)",
       fill = "Variable Type") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

The random forest correctly assigns the highest importance to the informative
predictors V1–V5, with V1 and V4 leading. The uninformative predictors V6–V10
receive very low importance, and notably V8, V9, and V10 received **negative**
importance scores (-0.10, -0.08, -0.05). In random forest importance, a negative
value means the variable performed worse than a random permutation, confirming
it is truly uninformative. Unlike MARS, random forests do not drive noise variable
importance to absolute zero through explicit pruning — instead, noise variables
occasionally get selected for random splits and may receive small spurious scores.
Overall, however, the model does not significantly rely on V6–V10.

## (b) Now add an additional predictor that is highly correlated with one of the informative predictors. For example:

> simulated$duplicate1 <- simulated$V1 + rnorm(200) * .1
> cor(simulated$duplicate1, simulated$V1)

## Fit another random forest model to these data. Did the importance score for V1 change? What happens when you add another predictor that is also highly correlated with V1?

```{r rf-duplicate, cache=TRUE}
# duplicate highly correlated with V1
simulated$duplicate1 <- simulated$V1 + rnorm(200) * 0.1
cat("Correlation between duplicate1 and V1:",
    round(cor(simulated$duplicate1, simulated$V1), 4), "\n")

set.seed(100)
model2 <- randomForest(y ~ ., data = simulated,
                       importance = TRUE,
                       ntree = 1000)
rfImp2 <- varImp(model2, scale = FALSE)

# add a second duplicate
simulated$duplicate2 <- simulated$V1 + rnorm(200) * 0.1
cat("Correlation between duplicate2 and V1:",
    round(cor(simulated$duplicate2, simulated$V1), 4), "\n")

set.seed(100)
model3 <- randomForest(y ~ ., data = simulated,
                       importance = TRUE,
                       ntree = 1000)
rfImp3 <- varImp(model3, scale = FALSE)

# V1 importance across the three models
cat("\nV1 importance — original model:   ", round(rfImp1["V1","Overall"], 2), "\n")
cat("V1 importance — one duplicate:    ", round(rfImp2["V1","Overall"], 2), "\n")
cat("V1 importance — two duplicates:   ", round(rfImp3["V1","Overall"], 2), "\n")
```

When a highly correlated duplicate of V1 is added, the importance score for V1
decreases noticeably. This is because the random forest now has two near-identical
predictors competing for the same splits — the signal that V1 carried gets shared
between V1 and its duplicate, so neither alone appears as important as V1 did
in the first place. Adding a second duplicate dilutes V1's importance further. This is not
surprising since it is known that the traditional random forest importance measure is
not stable in the presence of correlated predictors, tending to spread importance
across correlated variables rather than concentrating it on the most informative one.

```{r cleanup-duplicates}
simulated$duplicate1 <- NULL
simulated$duplicate2 <- NULL
```

## (c) Use the cforest function in the party package to fit a random forest model using conditional inference trees. The party package function varimp can calculate predictor importance. The conditional argument of that function toggles between the traditional importance measure and the modified version described in Strobl et al. (2007). Do these importances show the same pattern as the traditional random forest model?

```{r cforest, cache=TRUE}
set.seed(100)
cfModel <- cforest(y ~ ., data = simulated,
                   controls = cforest_unbiased(ntree = 1000, mtry = 5))

# traditional importance
cfImp_trad <- varimp(cfModel, conditional = FALSE)
# conditional importance 
cfImp_cond <- varimp(cfModel, conditional = TRUE)

imp_compare <- data.frame(
  Predictor   = names(cfImp_trad),
  Traditional = round(cfImp_trad, 4),
  Conditional = round(cfImp_cond, 4),
  Informative = names(cfImp_trad) %in% paste0("V", 1:5)
)
imp_compare <- imp_compare[order(-imp_compare$Traditional), ]
print(imp_compare)
```

The conditional importance measure (Strobl et al. 2007) corrects for the bias
that the traditional measure exhibits when predictors are correlated. With the
conditional approach, uninformative predictors V6–V10 are pushed closer to zero,
while the ranking of informative predictors V1–V5 is preserved. This is the same
general pattern as the standard random forest, but the conditional importance is
more conservative and trustworthy when predictors share information.

## (d)Repeat this process with different tree models, such as boosted trees and Cubist. Does the same pattern occur?

```{r gbm-model, cache=TRUE}
set.seed(100)
gbmGrid <- expand.grid(
  interaction.depth = c(1, 3, 5),
  n.trees           = c(100, 500, 1000),
  shrinkage         = 0.1,
  n.minobsinnode    = 10
)
gbmMod <- train(y ~ ., data = simulated,
                method    = "gbm",
                tuneGrid  = gbmGrid,
                trControl = trainControl(method = "cv", number = 5),
                verbose   = FALSE)

gbmImp <- varImp(gbmMod, scale = TRUE)
print(gbmImp, top = 10)
```

```{r cubist-model, cache=TRUE}
set.seed(100)
cubistMod <- train(y ~ ., data = simulated,
                   method    = "cubist",
                   trControl = trainControl(method = "cv", number = 5))

cubistImp <- varImp(cubistMod, scale = TRUE)
print(cubistImp, top = 10)
```

Both gradient boosting and Cubist show the same broad pattern: informative
predictors V1–V5 receive the highest importance, while V6–V10 receive near-zero
scores. Gradient boosting is particularly good at suppressing noise variables because
each tree is built to correct the residuals of the previous one, meaning uninformative
splits quickly contribute nothing to the gradient. Cubist, being a rule-based model,
also tends to exclude irrelevant predictors from its rules. However, as with the
standard random forest, neither model drives noise variable importance to absolute
zero as cleanly as MARS does — confirming that GCV-based pruning remains the
most aggressive form of built-in variable selection among these methods.

One notable difference from MARS: in the GBM output, **V3 received an importance
of 28.48%**, substantially higher than the near-zero score MARS assigned it. This
reflects that GBM is less selective and picks up weaker signals more readily. V3
does appear in the true Friedman equation as $20(x_3-0.5)^2$, so some importance
is legitimate, but GBM surfaces it more readily than the GCV-pruned MARS model.

---

# Question 8.2: Use a simulation to show tree bias with different granularities.

This exercise asks to demonstrate through simulation that regression trees
exhibit selection bias toward predictors with more unique values (higher
granularity). A tree will prefer to split on a continuous variable over a binary
one, even when both are equally uninformative, simply because more split points
are available.

## Simulation

```{r tree-bias-sim}
set.seed(100)
n <- 500

x_continuous <- runif(n, 0, 1)

x_binary <- sample(c(0, 1), n, replace = TRUE)

y_noise <- rnorm(n)

bias_df <- data.frame(y = y_noise,
                      x_continuous = x_continuous,
                      x_binary     = x_binary)
```

```{r tree-bias-fit}
# Fit a single regression tree
set.seed(100)
bias_tree <- rpart(y ~ x_continuous + x_binary,
                   data    = bias_df,
                   control = rpart.control(cp = 0.001, minsplit = 10))

# Variable importance from tree
vi_bias <- bias_tree$variable.importance
cat("Tree variable importance:\n")
print(round(vi_bias, 4))
```

```{r tree-bias-plot, fig.height=4}
vi_df_bias <- data.frame(
  Predictor  = names(vi_bias),
  Importance = as.numeric(vi_bias)
)

ggplot(vi_df_bias, aes(x = reorder(Predictor, Importance),
                       y = Importance, fill = Predictor)) +
  geom_col(width = 0.5, color = "white") +
  coord_flip() +
  scale_fill_manual(values = c("x_continuous" = "#5b8dd9",
                               "x_binary"     = "#e07b54")) +
  labs(title    = "Q8.2 – Tree Bias: Continuous vs Binary Predictor",
       subtitle = "Both predictors are uninformative (pure noise response)\nbut the tree favors the continuous predictor due to more available split points",
       x = "Predictor", y = "Importance") +
  theme_minimal(base_size = 13) +
  theme(legend.position  = "none",
        plot.title       = element_text(face = "bold"))
```

Even though both predictors are completely uninformative (the response is
random noise), the regression tree assigns higher importance to the continuous
predictor than to the binary one — specifically, 199.93 vs 16.85, a roughly
**12x difference**. This happens because a continuous variable offers many more
candidate split points, making it more likely that at least one split will appear
to reduce node impurity by chance. A binary predictor has only one possible split
point, so it has far fewer opportunities to appear useful. This is the granularity
bias of regression trees: they systematically prefer high-cardinality predictors,
which can lead to misleading variable importance rankings when predictors differ
in the number of unique values they take.

---

# Question 8.3 
In stochastic gradient boosting the bagging fraction and learning rate will govern the construction of the trees as they are guided by the gradient. Although the optimal values of these parameters should be obtained through the tuning process, it is helpful to understand how the magnitudes of these parameters affect magnitudes of variable importance. Figure 8.24 provides the variable importance plots for boosting using two extreme values for the bagging fraction (0.1 and 0.9) and the learning rate (0.1 and 0.9) for the solubility data. The left-hand plot has both parameters set to 0.1, and the right-hand plot has both set to 0.9:


This question is asking to interpret Figure 8.24 from the book, which shows
variable importance plots for two extreme boosting configurations on the
solubility data:

- **Left plot:** bagging fraction = 0.1, learning rate = 0.1 (both low)
- **Right plot:** bagging fraction = 0.9, learning rate = 0.9 (both high)

## (a)Why does the model on the right focus its importance on just the first few of predictors, whereas the model on the left spreads importance across more predictors?

When both the bagging fraction and learning rate are set to 0.9, each tree
uses nearly the entire dataset (90%) and takes large gradient steps. This means
the model converges quickly and aggressively, meaning that a small number of the strongest
predictors dominate from the very first trees, and subsequent trees continue to
exploit the same strong signals. The large learning rate also means that the
residuals are reduced rapidly, leaving little room for weaker predictors to
contribute. The result is a steep importance curve concentrated on just a few
variables.

However, when both parameters are set to 0.1, each tree sees only 10% of
the data (introducing high variance through subsampling) and takes tiny gradient
steps. This forces the model to explore a broader range of predictors over many
more trees, spreading importance more evenly across the predictor set. The
subsampling also acts as a regularizer, preventing any single predictor from
dominating too early.

## (b) Which model do you think would be more predictive of other samples?


The left model (both parameters = 0.1) is likely to generalize better to
new samples. The combination of a small bagging fraction and a low learning rate
acts as strong regularization — the subsampling introduces randomness that
prevents overfitting, and the small step size means the model improves gradually
and carefully. The right model, with its aggressive learning rate and large bagging
fraction, risks overfitting to the training data by quickly memorizing the strongest
patterns. In practice, the optimal values should always be determined through
cross-validation, but between these two extremes, the slower, more exploratory
left model is the safer choice for out-of-sample prediction.

## (c)  How would increasing interaction depth affect the slope of predictor importance for either model in Fig. 8.24?

Increasing the interaction depth allows individual trees to model higher-order
interactions between predictors. This generally flattens the importance curve
(makes the slope less steep) because deeper trees can capture complex joint
effects that involve more predictors simultaneously. When interactions are modeled,
predictors that were previously uninformative on their own can become important
in combination with others, spreading importance more broadly across the predictor
set. This effect is more pronounced in the right model (high bagging fraction,
high learning rate), where the current steep drop-off in importance would be
softened as more predictors get involved in deeper interaction splits. In the
left model, the importance is already spread out, so the effect of increasing
depth would be less dramatic.

---

# Question 8.7 Refer to Exercises 6.3 and 7.5 which describe a chemical manufacturing process. Use the same data imputation, data splitting, and pre-processing steps as before and train several tree-based models:


Using the same `ChemicalManufacturingProcess` dataset and pre-processing
pipeline from Exercises 6.3 and 7.5 (median imputation, near-zero variance
removal, centering and scaling), now I will train several tree-based regression models.

## Pre-processing & Data Splitting

```{r load-chem}
data(ChemicalManufacturingProcess)
```

```{r preprocess-87}
set.seed(42)
trainIdx <- createDataPartition(ChemicalManufacturingProcess$Yield,
                                p = 0.8, list = FALSE)
trainRaw <- ChemicalManufacturingProcess[ trainIdx, ]
testRaw  <- ChemicalManufacturingProcess[-trainIdx, ]

preProcVals <- preProcess(trainRaw[, -1],
                          method = c("medianImpute", "nzv",
                                     "center", "scale"))
trainX <- predict(preProcVals, trainRaw[, -1])
testX  <- predict(preProcVals, testRaw[, -1])
trainY <- trainRaw$Yield
testY  <- testRaw$Yield

cat("Train:", nrow(trainX), "| Test:", nrow(testX),
    "| Predictors after preprocessing:", ncol(trainX), "\n")

ctrl10 <- trainControl(method = "cv", number = 10)
```

## (a) Which tree-based regression model gives the optimal resampling and test set performance?

### Single Regression Tree (rpart)

```{r rpart-87, cache=TRUE}
set.seed(100)
rpartMod <- train(x = trainX, y = trainY,
                  method    = "rpart",
                  tuneLength = 20,
                  trControl = ctrl10)
rpartPred <- predict(rpartMod, testX)
rpart_ts  <- postResample(rpartPred, testY)
cat("Single Tree — CV RMSE:", round(min(rpartMod$results$RMSE), 4),
    "| Test RMSE:", round(rpart_ts["RMSE"], 4),
    "| Test R²:", round(rpart_ts["Rsquared"], 4), "\n")
```

### Random Forest

```{r rf-87, cache=TRUE}
set.seed(100)
rfMod <- train(x = trainX, y = trainY,
               method    = "rf",
               tuneLength = 5,
               trControl = ctrl10,
               ntree     = 500)
rfPred <- predict(rfMod, testX)
rf_ts  <- postResample(rfPred, testY)
cat("Random Forest — CV RMSE:", round(min(rfMod$results$RMSE), 4),
    "| Test RMSE:", round(rf_ts["RMSE"], 4),
    "| Test R²:", round(rf_ts["Rsquared"], 4), "\n")
```

### Gradient Boosting (GBM)

```{r gbm-87, cache=TRUE}
set.seed(100)
gbmGrid87 <- expand.grid(
  interaction.depth = c(1, 3, 5),
  n.trees           = c(50, 100, 300),
  shrinkage         = c(0.01, 0.1),
  n.minobsinnode    = 10
)
gbmMod87 <- train(x = trainX, y = trainY,
                  method    = "gbm",
                  tuneGrid  = gbmGrid87,
                  trControl = ctrl10,
                  verbose   = FALSE)
gbmPred87 <- predict(gbmMod87, testX)
gbm_ts87  <- postResample(gbmPred87, testY)
cat("GBM — CV RMSE:", round(min(gbmMod87$results$RMSE), 4),
    "| Test RMSE:", round(gbm_ts87["RMSE"], 4),
    "| Test R²:", round(gbm_ts87["Rsquared"], 4), "\n")
```

### Cubist

```{r cubist-87, cache=TRUE}
set.seed(100)
cubistMod87 <- train(x = trainX, y = trainY,
                     method    = "cubist",
                     trControl = ctrl10)
cubistPred87 <- predict(cubistMod87, testX)
cubist_ts87  <- postResample(cubistPred87, testY)
cat("Cubist — CV RMSE:", round(min(cubistMod87$results$RMSE), 4),
    "| Test RMSE:", round(cubist_ts87["RMSE"], 4),
    "| Test R²:", round(cubist_ts87["Rsquared"], 4), "\n")
```

### Performance Summary

```{r summary-87}
summ87 <- data.frame(
  Model     = c("Single Tree", "Random Forest", "GBM", "Cubist"),
  CV_RMSE   = round(c(min(rpartMod$results$RMSE),
                      min(rfMod$results$RMSE),
                      min(gbmMod87$results$RMSE),
                      min(cubistMod87$results$RMSE)), 4),
  Test_RMSE = round(c(rpart_ts["RMSE"], rf_ts["RMSE"],
                      gbm_ts87["RMSE"], cubist_ts87["RMSE"]), 4),
  Test_Rsq  = round(c(rpart_ts["Rsquared"], rf_ts["Rsquared"],
                      gbm_ts87["Rsquared"], cubist_ts87["Rsquared"]), 4)
)
print(summ87)
```

```{r plot-perf-87}
summ87_long <- pivot_longer(summ87, cols = c(CV_RMSE, Test_RMSE),
                             names_to = "Set", values_to = "RMSE")
ggplot(summ87_long, aes(x = reorder(Model, -RMSE), y = RMSE, fill = Set)) +
  geom_col(position = "dodge", width = 0.6, color = "white") +
  geom_text(aes(label = round(RMSE, 2)),
            position = position_dodge(0.6), vjust = -0.3, size = 3.5) +
  scale_fill_manual(values = c("CV_RMSE"   = "#5b8dd9",
                               "Test_RMSE" = "#e07b54"),
                    labels  = c("10-Fold CV RMSE", "Test RMSE")) +
  labs(title    = "Q8.7(a) – Tree-Based Model Performance: Chemical Mfg. Process",
       x = "Model", y = "RMSE", fill = "") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

**Cubist is the best-performing tree-based model**, achieving the lowest CV RMSE
(1.00) and the lowest test RMSE (0.95) along with the highest R² (0.795). GBM
is a close second on CV (1.06) but its test RMSE (1.16) is notably worse than
Cubist, suggesting Cubist generalizes better on this dataset. Random Forest
performs reasonably well (test RMSE = 1.18, R² = 0.733), while the single
regression tree is by far the weakest model (test RMSE = 1.49, R² = 0.463),
which is expected — a single tree is high variance and prone to overfitting
compared to ensemble methods. The clear ranking is: Cubist > GBM > Random
Forest > Single Tree.

## (b)  Which predictors are most important in the optimal tree-based regression model? Do either the biological or process variables dominate the list? How do the top 10 important predictors compare to the top 10 predictors from the optimal linear and nonlinear models?

```{r varimp-87}
# best model for variable importance
vi_best <- varImp(cubistMod87, scale = TRUE)
print(vi_best, top = 15)
```

```{r plot-varimp-87}
vi_df87 <- vi_best$importance
vi_df87$Predictor <- rownames(vi_df87)
vi_df87 <- vi_df87[order(-vi_df87$Overall), ]
vi_df87$Type <- ifelse(grepl("Bio", vi_df87$Predictor),
                       "Biological", "ManufacturingProcess")
vi_df87_clean <- vi_df87[!is.na(vi_df87$Overall), ]

ggplot(head(vi_df87_clean, 10),
       aes(x = reorder(Predictor, Overall), y = Overall, fill = Type)) +
  geom_col(width = 0.6, color = "white") +
  coord_flip() +
  scale_fill_manual(values = c("Biological"           = "#4CAF50",
                               "ManufacturingProcess" = "#5b8dd9")) +
  labs(title    = "Q8.7(b) – Top Variable Importance (Best Tree Model)",
       x = "Predictor", y = "Relative Importance (%)",
       fill = "Variable Type") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

Manufacturing process variables dominate the top of the list, with
ManufacturingProcess32 (100%), ManufacturingProcess09 (66.67%), and
ManufacturingProcess17 (57.14%) leading. However, unlike the MARS model from
Q7.5 — which selected exclusively manufacturing process variables — the Cubist
model surfaces **two biological material variables** in the top 10:
BiologicalMaterial06 (36.19%, rank 4) and BiologicalMaterial03 (22.86%, rank 10).
This broader mix suggests Cubist's rule-based structure is capturing a richer
set of predictors than MARS's sparse hinge functions. Compared to the Chapter 6
linear model, which also typically shows a mix of both variable types, the
Cubist top 10 shares several of the same process variables (ManufacturingProcess32,
ManufacturingProcess09, ManufacturingProcess17) but additionally includes
process variables like ManufacturingProcess29 and ManufacturingProcess45 that
did not appear in the MARS nonlinear model, reflecting the different ways each
model partitions the predictor space.

## (c) Optimal Single Tree Plot

```{r single-tree-plot, fig.width=12, fig.height=7}
# Refit a pruned single tree for visualization
set.seed(100)
final_tree <- rpart(Yield ~ ., 
                    data    = ChemicalManufacturingProcess,
                    control = rpart.control(cp = rpartMod$bestTune$cp))

rpart.plot(final_tree,
           type   = 4,
           extra  = 101,
           fallen.leaves = TRUE,
           main   = "Q8.7(c) – Optimal Single Regression Tree\nChemical Manufacturing Process → Yield",
           cex    = 0.7)
```

The single tree visualization reveals the most important binary decision rules
the model uses to partition yield. The **root split on ManufacturingProcess32 < 160**
is consistent with it being the top predictor in part (b), cleanly dividing the
176 observations into a lower-yield group (n=103, mean yield ≈ 39) and a
higher-yield group (n=73). Notably, biological material variables appear at
**deeper levels** of the tree — BiologicalMaterial11, BiologicalMaterial05, and
BiologicalMaterial02 all appear as secondary splits — suggesting they matter
conditionally, only within specific process operating ranges. This is an
important insight: the manufacturing process sets the primary yield ceiling,
while biological material quality provides finer differentiation within those
process-defined groups. The terminal nodes range from a predicted yield of 37
(n=8, 5%) up to 42 (n=8, 5%), showing that the combination of process and
biological conditions creates meaningful yield variation. While a single tree
is less accurate than ensemble methods, each path from root to leaf is a
readable set of if-then rules that provides clear operational intuition.

