---
title: "Data 624 Homework 7"
author: "Joao De Oliveira"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

## Introduction
In this report, I will be answering questions 2 and 3 in the 6th chapter
from the Kuhn and Johnson book.


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 7,
  fig.height = 5,
  fig.align  = "center"
)
```


## Libraries

```{r}
library(AppliedPredictiveModeling)
library(caret)
library(pls)
library(glmnet)
library(MASS)
library(dplyr)
library(ggplot2)
library(tidyverse)
```


### Question 6.2 
Developing a model to predict permeability (see Sect. 1.4) could save significant resources for a pharmaceutical company, while at the same time more rapidly identifying molecules that have a sufficient permeability to become a drug:

#### (a) 
Start R and use these commands to load the data:
> library(AppliedPredictiveModeling)
> data(permeability)

The matrix fingerprints contains the 1,107 binary molecular predictors for the 165 compounds, while permeability contains permeability response.


```{r}
set.seed(1234)

# data
data(permeability)

# dimensions of the data
fingerprint_dim <- dim(fingerprints)
response_length <- length(permeability)

fingerprint_dim
response_length
```

```{r inspect-data}
str(fingerprints)
summary(permeability)
```

`fingerprints` is a 165 x 1107 binary matrix where each column indicates whether a molecular substructure is absent or present. The object `permeability` is a continuous response vector of length 165. 

The response variable appears to be right-skewed because the mean is much larger than the median and the maximum value is far above the third quartile. Because of this, I considered two versions of the response: the original permeability values and a log-transformed version.

Response distribution

```{r}
response_df <- data.frame(
  permeability = permeability,
  log_permeability = log(permeability)
)

summary(response_df)

response_df |>
  pivot_longer(cols = everything(), names_to = "Response", values_to = "Value") |>
  ggplot(aes(x = Value)) +
  geom_histogram(bins = 25, color = "white") +
  facet_wrap(~ Response, scales = "free") +
  labs(
    title = "Distribution of Original and Log-Transformed Permeability",
    x = "Response value",
    y = "Count"
  ) +
  theme_minimal()
```


#### (b) 
The fingerprint predictors indicate the presence or absence of substructures of a molecule and are often sparse meaning that relatively few of the molecules contain each substructure. Filter out the predictors that have low frequencies using the nearZeroVar function from the caret package. How many predictors are left for modeling?


```{r near-zero-var}
nzv_index <- nearZeroVar(fingerprints)

fingerprints_filtered <- fingerprints[, -nzv_index]

original_predictors <- ncol(fingerprints)
removed_predictors <- length(nzv_index)
remaining_predictors <- ncol(fingerprints_filtered)

original_predictors
removed_predictors
remaining_predictors
```

After filtering out near-zero variance predictors, `r remaining_predictors` predictors are left for modeling.This step substantially reduces dimensionality while retaining the most informative molecular features, which is especially important given the sparsity of the fingerprint data.


#### (c)
Split the data into a training and a test set, pre-process the data, and
tune a PLS model. How many latent variables are optimal and what is
the corresponding resampled estimate of R2?

I split the data (original and log-transformed) into a training set and a test set using `createDataPartition()` so that the distribution of the response variable is represented in both sets. 

```{r train-test-split}
set.seed(1234)
train_index <- createDataPartition(permeability, p = 0.80, list = FALSE)

x_train <- fingerprints_filtered[train_index, ]
x_test  <- fingerprints_filtered[-train_index, ]

y_train <- permeability[train_index]
y_test  <- permeability[-train_index]

y_train_log <- log(y_train)
y_test_log  <- log(y_test)

nrow(x_train)
nrow(x_test)
```

Because the predictors are binary, centering and scaling are not always required in the same way as for continuous predictors. However, for PLS and other regression models, using consistent preprocessing is helpful. Here, I will center and scale the predictors inside the `train()` function.

Train Control

```{r}
set.seed(1234)
ctrl <- trainControl(
  method = "repeatedcv",
  number = 10,
  repeats = 5
)

pls_grid <- expand.grid(ncomp = 1:20)
```


**PLS on the Original Response**

```{r pls-original-model}
set.seed(1234)
pls_fit <- train(
  x = x_train,
  y = y_train,
  method = "pls",
  preProcess = c("center", "scale"),
  tuneGrid = pls_grid,
  trControl = ctrl,
  metric = "Rsquared"
)

pls_fit
plot(pls_fit)
```

```{r pls-original-best-results}
best_pls_ncomp <- pls_fit$bestTune$ncomp
best_pls_results <- pls_fit$results |>
  filter(ncomp == best_pls_ncomp)

best_pls_ncomp
best_pls_results
```

**PLS on the Log-Transformed Response**

```{r pls-log-model}
set.seed(1234)
pls_log_fit <- train(
  x = x_train,
  y = y_train_log,
  method = "pls",
  preProcess = c("center", "scale"),
  tuneGrid = pls_grid,
  trControl = ctrl,
  metric = "Rsquared"
)

pls_log_fit
plot(pls_log_fit)
```

```{r pls-log-best-results}
best_pls_log_ncomp <- pls_log_fit$bestTune$ncomp
best_pls_log_results <- pls_log_fit$results |>
  filter(ncomp == best_pls_log_ncomp)

best_pls_log_ncomp
best_pls_log_results
```

For the PLS model using the original response, the optimal number of latent variables was 7, with a cross-validated R² of approximately 0.57.

For the log-transformed response, the optimal number of components increased to 9, with a cross-validated R² of approximately 0.61 (on the log scale).

The slightly higher R² for the log-transformed response suggests that reducing skewness improves the model’s ability to capture underlying structure. However, these values are not directly comparable across scales and should be interpreted cautiously.


#### (d)
Predict the response for the test set. What is the test set estimate of R2?

For the original-response model, predictions are already on the original permeability scale. For the log-response model, predictions are first made on the log scale and then back-transformed with `exp()` so that test-set performance can also be evaluated on the original permeability scale.


```{r metric-functions}
get_metrics <- function(obs, pred, model_name, response_scale) {
  data.frame(
    Model = model_name,
    Response_Scale = response_scale,
    Test_R2 = cor(obs, pred)^2,
    Test_RMSE = RMSE(pred, obs),
    Test_MAE = MAE(pred, obs)
  )
}
```

```{r pls-test-predictions}
# original-response PLS
pls_pred <- predict(pls_fit, newdata = x_test)

# log-response PLS
pls_log_pred_log_scale <- predict(pls_log_fit, newdata = x_test)
pls_log_pred_original_scale <- exp(pls_log_pred_log_scale)

pls_test_results <- bind_rows(
  get_metrics(y_test, pls_pred, "PLS", "Original response"),
  get_metrics(y_test_log, pls_log_pred_log_scale, "PLS", "Log response / log-scale test"),
  get_metrics(y_test, pls_log_pred_original_scale, "PLS", "Log response / back-transformed test")
)

pls_test_results
```

Test set performance provides a more realistic estimate of model generalization. When evaluated on the original permeability scale, the log-transformed PLS model (after back-transformation) achieved a higher R² (~0.31) than the model trained on the original response (~0.22).

This indicates that transforming the response improved predictive performance in practical terms, even after converting predictions back to the original scale. However, the overall R² values remain relatively low, suggesting that a substantial portion of variability is not captured by the model.

PLS Observed vs Predicted

```{r}
pls_plot_data <- bind_rows(
  data.frame(
    Observed = y_test,
    Predicted = pls_pred,
    Model = "PLS: Original response"
  ),
  data.frame(
    Observed = y_test,
    Predicted = pls_log_pred_original_scale,
    Model = "PLS: Log response, back-transformed"
  )
)

ggplot(pls_plot_data, aes(x = Observed, y = Predicted)) +
  geom_point(alpha = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  facet_wrap(~ Model) +
  labs(
    title = "PLS Models: Observed vs Predicted Permeability",
    x = "Observed Permeability",
    y = "Predicted Permeability"
  ) +
  theme_minimal()
```


#### (e)
Try building other models discussed in this chapter. Do any have better
predictive performance?


I will compare PLS with ridge regression, lasso regression, elastic net, and principal component regression. Each model is fit twice: once using the original response and once using the log-transformed response.


Model Grids

```{r}
ridge_grid <- expand.grid(
  alpha = 0,
  lambda = 10^seq(-4, 2, length = 50)
)

lasso_grid <- expand.grid(
  alpha = 1,
  lambda = 10^seq(-4, 2, length = 50)
)

enet_grid <- expand.grid(
  alpha = seq(0.1, 0.9, by = 0.2),
  lambda = 10^seq(-4, 2, length = 30)
)

pcr_grid <- expand.grid(ncomp = 1:20)
```

Helper Function to Train both Modelts

```{r train-compare-helper}
train_original_and_log <- function(model_label, method_name, tune_grid) {
  set.seed(1234)
  fit_original <- train(
    x = x_train,
    y = y_train,
    method = method_name,
    preProcess = c("center", "scale"),
    tuneGrid = tune_grid,
    trControl = ctrl,
    metric = "Rsquared"
  )
  
  set.seed(1234)
  fit_log <- train(
    x = x_train,
    y = y_train_log,
    method = method_name,
    preProcess = c("center", "scale"),
    tuneGrid = tune_grid,
    trControl = ctrl,
    metric = "Rsquared"
  )
  
  pred_original <- predict(fit_original, newdata = x_test)
  pred_log_scale <- predict(fit_log, newdata = x_test)
  pred_log_backtransformed <- exp(pred_log_scale)
  
  test_results <- bind_rows(
    get_metrics(y_test, pred_original, model_label, "Original response"),
    get_metrics(y_test_log, pred_log_scale, model_label, "Log response / log-scale test"),
    get_metrics(y_test, pred_log_backtransformed, model_label, "Log response / back-transformed test")
  )
  
  cv_results <- bind_rows(
    fit_original$results |>
      filter(Rsquared == max(Rsquared)) |>
      slice(1) |>
      mutate(Model = model_label, Response_Scale = "Original response"),
    fit_log$results |>
      filter(Rsquared == max(Rsquared)) |>
      slice(1) |>
      mutate(Model = model_label, Response_Scale = "Log response")
  ) |>
    select(Model, Response_Scale, everything())
  
  list(
    fit_original = fit_original,
    fit_log = fit_log,
    test_results = test_results,
    cv_results = cv_results
  )
}
```

**Ridge Regression**

```{r ridge-models}
ridge_models <- train_original_and_log("Ridge Regression", "glmnet", ridge_grid)

ridge_models$fit_original$bestTune
ridge_models$fit_log$bestTune
ridge_models$test_results
```

**Lasso Regression**

```{r lasso-models}
lasso_models <- train_original_and_log("Lasso Regression", "glmnet", lasso_grid)

lasso_models$fit_original$bestTune
lasso_models$fit_log$bestTune
lasso_models$test_results
```

**Elastic Net**

```{r elastic-net-models}
enet_models <- train_original_and_log("Elastic Net", "glmnet", enet_grid)

enet_models$fit_original$bestTune
enet_models$fit_log$bestTune
enet_models$test_results
```

**Principal Component Regression**

```{r pcr-models}
pcr_models <- train_original_and_log("PCR", "pcr", pcr_grid)

pcr_models$fit_original$bestTune
pcr_models$fit_log$bestTune
pcr_models$test_results
plot(pcr_models$fit_original)
plot(pcr_models$fit_log)
```

**Model Comparison**

```{r model-comparison}
all_test_results <- bind_rows(
  pls_test_results,
  ridge_models$test_results,
  lasso_models$test_results,
  enet_models$test_results,
  pcr_models$test_results
)

# predictions evaluated on original permeability scale
original_scale_comparison <- all_test_results |>
  filter(Response_Scale %in% c("Original response", "Log response / back-transformed test")) |>
  arrange(desc(Test_R2), Test_RMSE)

original_scale_comparison
```

```{r model-comparison-plot}
ggplot(original_scale_comparison, aes(x = reorder(paste(Model, Response_Scale, sep = " - "), Test_R2), y = Test_R2)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Comparison of Test Set R-squared Values on the Original Permeability Scale",
    x = "Model and response version",
    y = "Test Set R-squared"
  ) +
  theme_minimal()
```

```{r cross-validation-comparison}
cv_comparison <- bind_rows(
  best_pls_results |> mutate(Model = "PLS", Response_Scale = "Original response"),
  best_pls_log_results |> mutate(Model = "PLS", Response_Scale = "Log response"),
  ridge_models$cv_results,
  lasso_models$cv_results,
  enet_models$cv_results,
  pcr_models$cv_results
) |>
  select(Model, Response_Scale, Rsquared, RMSE, MAE, everything()) |>
  arrange(desc(Rsquared))

cv_comparison
```

The table above compares the best resampled/cross-validated results. It is important to remember that the original-response and log-response cross-validated \(R^2\) values are calculated on different response scales, so they should not be interpreted as perfectly equivalent. For the final practical comparison, the test-set results on the original permeability scale are the most useful.

Among all models evaluated, the lasso regression model trained on the log-transformed response (with predictions back-transformed to the original scale) achieved the best performance, with a test R² of approximately 0.37. Elastic net and PLS models followed closely, while ridge regression and PCR performed worse overall.

The relatively small differences between the top models suggest that, while regularization improves performance, the primary limitation lies in the data representation and underlying complexity of the problem rather than the choice of model alone.

#### (f)
Would you recommend any of your models to replace the permeability
laboratory experiment?

Although some models achieved moderate predictive performance, the best test R² was approximately 0.37, indicating that a large portion of the variability in permeability remains unexplained. In a pharmaceutical setting, where accurate predictions are critical for decision-making, this level of performance is likely insufficient to replace laboratory experiments. However, these models could still be useful as a screening tool to prioritize compounds before experimental testing. The log transformation consistently improved model performance on the original scale, suggesting that addressing skewness was beneficial. Even with advanced techniques and regularization, predictive performance may remain limited when the signal-to-noise ratio is low or important nonlinear relationships are not captured.


### Question 6.3
A chemical manufacturing process for a pharmaceutical product was
discussed in Sect. 1.4. In this problem, the objective is to understand the relationship between biological measurements of the raw materials (predictors), 6.5 Computing 139 measurements of the manufacturing process (predictors), and the response of product yield. Biological predictors cannot be changed but can be used to assess the quality of the raw material before processing. On the other hand, manufacturing process predictors can be changed in the manufacturing process. Improving product yield by 1 % will boost revenue by approximately one hundred thousand dollars per batch:

#### (a)
Start R and use these commands to load the data:
> library(AppliedPredictiveModeling)
> data(chemicalManufacturing)

The matrix processPredictors contains the 57 predictors (12 describing
the input biological material and 45 describing the process predictors)
for the 176 manufacturing runs. yield contains the percent yield for each
run.


```{r chemical-load-data}
data(ChemicalManufacturingProcess)

chem_df <- ChemicalManufacturingProcess

# split response and predictors
chem_y <- chem_df$Yield
chem_x <- chem_df[,-1]

chem_dim <- dim(chem_x)
yield_length <- length(chem_y)

chem_dim
yield_length

str(chem_x)
summary(chem_y)
```

`ChemicalManufacturingProcess` contains 176 manufacturing runs. The response variable is `Yield`, and the remaining 57 columns are predictors. The first 12 predictors describe the biological raw material, while the remaining 45 predictors describe the manufacturing process.

```{r chemical-predictor-types}
bio_predictors <- names(chem_x)[grepl("BiologicalMaterial", names(chem_x))]
process_predictors <- names(chem_x)[grepl("ManufacturingProcess", names(chem_x))]

length(bio_predictors)
length(process_predictors)

head(bio_predictors)
head(process_predictors)
```

#### (b)
A small percentage of cells in the predictor set contain missing values. Use an imputation function to fill in these missing values (e.g., see Sect. 3.8)


A small number of predictor values are missing. Since the models in this chapter require complete predictor data, I used KNN imputation. KNN imputation estimates missing values using similar observations, which makes sense here because the biological and manufacturing-process measurements are likely related to one another.


```{r chem-missing-values}
# total missing values 
sum(is.na(chem_x))

# predictors with at least one missing value
sum(colSums(is.na(chem_x)) > 0)

# Show the predictors with missing values
missing_summary <- data.frame(
  Predictor = names(chem_x),
  Missing_Count = colSums(is.na(chem_x))
) %>%
  filter(Missing_Count > 0) %>%
  arrange(desc(Missing_Count))

missing_summary
```

To show that KNN imputation removes the missing values, I also apply it once to the full predictor set for checking purposes. The final model below still performs the preprocessing inside `train()` so that the same steps are applied correctly during resampling and prediction.

```{r chemical-knn-imputation-check}
chem_knn_check <- preProcess(chem_x, method = c("center", "scale", "knnImpute"))
chem_x_knn_check <- predict(chem_knn_check, chem_x)

# confirm numb of missinf values
sum(is.na(chem_x_knn_check))
```

For the modeling step, KNN imputation is performed inside the `train()` function by including `knnImpute` in the `preProcess` argument.For each missing cell, the algorithm finds the *k* most similar complete observations (distance on non-missing features) and replaces the missing value with their weighted mean. This preserves local correlation structure better than simple mean/median imputation.


#### (c)
Split the data into a training and a test set, pre-process the data, and
tune a model of your choice from this chapter. What is the optimal value
of the performance metric?

I used an elastic net regression model. Elastic net is a useful choice for this problem because it combines ridge and lasso penalties. This allows it to handle correlated predictors while also shrinking less useful predictors toward zero.

```{r chemical-train-test-split}
set.seed(1234)
chem_train_index <- createDataPartition(chem_y, p = 0.80, list = FALSE)

chem_x_train <- chem_x[chem_train_index, ]
chem_x_test <- chem_x[-chem_train_index, ]
chem_y_train <- chem_y[chem_train_index]
chem_y_test <- chem_y[-chem_train_index]

nrow(chem_x_train)
nrow(chem_x_test)
```

```{r chemical-train-control}
set.seed(1234)
chem_ctrl <- trainControl(
  method = "repeatedcv",
  number = 10,
  repeats = 5
)

chem_enet_grid <- expand.grid(
  alpha = seq(0.1, 0.9, by = 0.2),
  lambda = 10^seq(-4, 2, length = 40)
)
```

```{r chemical-elastic-net-model}
set.seed(1234)
chem_enet_fit <- train(
  x = chem_x_train,
  y = chem_y_train,
  method = "glmnet",
  preProcess = c("center", "scale", "knnImpute"),
  tuneGrid = chem_enet_grid,
  trControl = chem_ctrl,
  metric = "Rsquared"
)

chem_enet_fit
plot(chem_enet_fit)
```

```{r chemical-best-tuning-results}
chem_best_tune <- chem_enet_fit$bestTune
chem_best_results <- chem_enet_fit$results %>%
  filter(alpha == chem_best_tune$alpha, lambda == chem_best_tune$lambda)

chem_best_tune
chem_best_results
```

The elastic net model achieved a maximum cross-validated R² of approximately 0.63. This indicates that the model is able to explain a substantial portion of the variability in yield during resampling. The optimal tuning parameters correspond to a relatively low-to-moderate value of alpha, suggesting that a combination of ridge and lasso regularization is beneficial for this dataset.


#### (d)
 Predict the response for the test set. What is the value of the performance metric and how does this compare with the resampled performance metric on the training set?
 
```{r chemical-test-performance}
chem_test_pred <- predict(chem_enet_fit, newdata = chem_x_test)

chem_test_results <- data.frame(
  Model = "Elastic Net",
  Test_R2 = cor(chem_y_test, chem_test_pred)^2,
  Test_RMSE = RMSE(chem_test_pred, chem_y_test),
  Test_MAE = MAE(chem_test_pred, chem_y_test)
)

chem_test_results
chem_best_results %>% select(alpha, lambda, RMSE, Rsquared, MAE)
```

The test set R² is lower than the cross-validated R², which is expected since resampling estimates are often optimistic. This difference suggests some degree of overfitting, although the gap is not excessively large. Overall, the model demonstrates reasonable generalization to unseen data.

```{r chemical-observed-predicted-plot}
chem_pred_df <- data.frame(
  Observed = chem_y_test,
  Predicted = chem_test_pred
)

ggplot(chem_pred_df, aes(x = Observed, y = Predicted)) +
  geom_point(alpha = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(
    title = "Elastic Net Model: Observed vs Predicted Yield",
    x = "Observed Yield",
    y = "Predicted Yield"
  ) +
  theme_minimal()
```


#### (e)
 Which predictors are most important in the model you have trained? Do
either the biological or process predictors dominate the list?


```{r chemical-variable-importance}
chem_varimp <- varImp(chem_enet_fit, scale = TRUE)
chem_varimp

chem_varimp_df <- chem_varimp$importance %>%
  tibble::rownames_to_column("Predictor") %>%
  arrange(desc(Overall)) %>%
  mutate(
    Predictor_Type = case_when(
      Predictor %in% bio_predictors ~ "Biological material",
      Predictor %in% process_predictors ~ "Manufacturing process",
      TRUE ~ "Other"
    )
  )

chem_top_predictors <- chem_varimp_df %>%
  slice_head(n = 15)

chem_top_predictors

chem_top_predictors %>%
  count(Predictor_Type)
```


The most important predictors are primarily manufacturing process variables rather than biological material variables. So, the yield is more strongly influenced by controllable process conditions than by the initial properties of the raw materials. While biological variables provide useful baseline information, the process variables appear to play a dominant role in determining final yield. So, operators can adjust those process parameters.

```{r chemical-variable-importance-plot}
ggplot(chem_top_predictors, aes(x = reorder(Predictor, Overall), y = Overall, fill = Predictor_Type)) +
  geom_col(show.legend = TRUE) +
  coord_flip() +
  labs(
    title = "Top Predictors of Manufacturing Yield",
    x = "Predictor",
    y = "Variable Importance"
  ) +
  theme_minimal()
```


#### (f)
Explore the relationships between each of the top predictors and the response. How could this information be helpful in improving yield in future runs of the manufacturing process?


To explore how the top predictors relate to yield, I plotted each of the top predictors against the response to help identify whether higher or lower values of a predictor tend to be associated with higher product yield.

```{r chemical-imputed-full-data}
chem_pp <- preProcess(chem_x, method = c("center", "scale", "knnImpute"))
chem_x_imputed_scaled <- predict(chem_pp, chem_x)

chem_plot_df <- chem_x_imputed_scaled %>%
  as.data.frame() %>%
  mutate(Yield = chem_y)
```

```{r chemical-top-predictor-relationships, fig.width=8, fig.height=6}
top_6_predictors <- chem_top_predictors$Predictor[1:6]

chem_plot_df %>%
  select(Yield, all_of(top_6_predictors)) %>%
  pivot_longer(cols = all_of(top_6_predictors), names_to = "Predictor", values_to = "Value") %>%
  ggplot(aes(x = Value, y = Yield)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ Predictor, scales = "free_x") +
  labs(
    title = "Relationships Between Top Predictors and Yield",
    x = "Predictor value (imputed, centered, and scaled)",
    y = "Yield"
  ) +
  theme_minimal()
```

The relationships between the top predictors and yield reveal how specific variables influence the manufacturing outcome. Several process variables show clear trends with yield, indicating that adjusting these parameters could lead to improved performance. This is particularly valuable because manufacturing process variables can be controlled. By identifying which variables have the strongest impact, engineers can optimize process settings to increase yield. Even small improvements are economically significant, as a 1% increase in yield corresponds to a substantial increase in revenue per batch. On another hand, biological material variables can't be directly controlled, but they can be used to assess the quality of incoming raw materials and anticipate potential yield issues before processing begins. The results highlight that while predictive modeling can explain a significant portion of yield variability, its greatest value lies in identifying actionable process variables that can be optimized to improve manufacturing outcomes.

