---
title: "DATA 624 Project 1: Time Series Forecasting"
author: "Joao De Oliveira"
date: "2026-03-28"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: paged
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = TRUE)
```


## Libraries

```{r}
library(dplyr)
library(tidyverse)
library(forecast)
library(writexl)
```
# Overview
This project consists of 3 parts:

**Part A – ATM Forecast, ATM624Data.xlsx**

In part A, I want you to forecast how much cash is taken out of 4 different ATM machines for May 2010.  The data is given in a single file.  The variable ‘Cash’ is provided in hundreds of dollars, other than that it is straight forward.   I am being somewhat ambiguous on purpose to make this have a little more business feeling.  Explain and demonstrate your process, techniques used and not used, and your actual forecast.  I am giving you data via an excel file, please provide your written report on your findings, visuals, discussion and your R code via an RPubs link along with the actual.rmd file  Also please submit the forecast which you will put in an Excel readable file.

**Part B – Forecasting Power, ResidentialCustomerForecastLoad-624.xlsx**

Part B consists of a simple dataset of residential power usage for January 1998 until December 2013.  Your assignment is to model these data and a monthly forecast for 2014.  The data is given in a single file.  The variable ‘KWH’ is power consumption in Kilowatt hours, the rest is straight forward.    Add this to your existing files above. 

**Part C – BONUS, optional (part or all), Waterflow_Pipe1.xlsx and Waterflow_Pipe2.xlsx**

Part C consists of two data sets.  These are simple 2 columns sets, however they have different time stamps.  Your optional assignment is to time-base sequence the data and aggregate based on hour (example of what this looks like, follows).  Note for multiple recordings within an hour, take the mean.  Then to determine if the data is stationary and can it be forecast.  If so, provide a week forward forecast and present results via Rpubs and .rmd and the forecast in an Excel readable file. 

# Part A: ATM Forecast

## Load and Clean Data

To ensure reproducibility, I converted the excel file into a CSV file and then 
uploaded to GitHub.

```{r}
atm_raw <- "https://raw.githubusercontent.com/JDO-MSDS/Data-624/refs/heads/main/Project%201/data/ATM624Data%20-%20ATM%20Data.csv"

atm <- read.csv(atm_raw)

glimpse(atm)
summary(atm)
```

```{r}
# fix date format since it was num
atm$DATE <- as.Date(atm$DATE, origin = "1899-12-30")

head(atm)
# checking how many different ATM
unique(atm$ATM)
```


```{r}
# removing the blank/missing ATM
atm <- atm |>
  filter(ATM != "")

unique(atm$ATM)
```

```{r}
# number of observations per ATM
atm |> count(ATM)

# split by atm
atm_list <- split(atm, atm$ATM)

# clean and sort
atm_list <- lapply(atm_list, function(df) {
  df |>
    filter(!is.na(Cash)) |>
    arrange(DATE)
})

# convert to time series
atm_ts <- lapply(atm_list, function(df) {
  ts_data <- ts(df$Cash, frequency = 7)
  na.interp(ts_data)
})
```
## ATM Plots

```{r}
par(mfrow = c(2,2)) 

for(name in names(atm_ts)) {
  plot(atm_ts[[name]],
       main = paste("Cash Withdrawals -", name),
       ylab = "Cash",
       xlab = "Time",
       col = "red")
}

par(mfrow = c(2,2))

for(name in names(atm_ts)) {
  acf(atm_ts[[name]],
      main = paste("ACF -", name))
}

par(mfrow = c(2,2))

for(name in names(atm_ts)) {
  pacf(atm_ts[[name]],
       main = paste("PACF -", name))
}
```


By looking at the plots, I can see that ATM1 and ATM2 show stable patterns over 
time, with somewhat regular ups and downs. The ACF plots suggest a repeating pattern
around every 7 days, which makes sense as a weekly pattern. The PACF plots also 
has some structure in the early lags, so it should be predictable. The weekly seasonality
and structured autocorrelation make ARIMA and ETS good model options. ATM3 is quite
different. Basically the only activity is near the end. This probably means that
this ATM was only installed by the end of the period. I will use a simple 
average-based approach due to the lack of data. ATM4 has a few extreme spikes. It
also has fluctuations. It is not easy to identify clear patterns, which can be
confirmed in the ACF and PACF plots that look more scattered. So, I will treat 
each ATM separately since they each show different behaviors. In the following 
section, I will implement these models and compare their forecasting performance.


## ATM1 Forecast

### Fit ARIMA model for ATM1

As explained above, the plot for ATM1 showed a weekly pattern, so I will use
`auto.arima()` to select the best model based on the data since seasonality makes
sense.

```{r}
atm1_arima <- auto.arima(atm_ts$ATM1, seasonal = TRUE)
summary(atm1_arima)
```
The model selected by `auto.arima()` is ARIMA(0,0,1)(0,1,1)[7]. The weekly seasonality
is confirmed.

### Check ARIMA diagnosis for ATM1

```{r}
checkresiduals(atm1_arima)
```

The residuals from the ARIMA(0,0,1)(0,1,1)[7] model fluctuate around zero with 
no obvious pattern, so that means that the model has captured the main structure
of the data. The ACF plot shows that correlations are within the confidence
limits, so no autocorrelation left. The Ljung-Box test produced a p-value of 0.5062,
so we fail to reject the null hypothesis of no autocorrelation. The residuals 
behave like white noise, telling us the model is a good fit.


### Fit ETS for ATM1

```{r}
atm1_ets <- ets(atm_ts$ATM1)
summary(atm1_ets)
```
The model shows a very small smoothing parameter for the level and some seasonal
adjustment, indicating that most of the variation comes from the weekly patterns.
The error measures (RMSE around 26.05, MAE 16.96) suggest a reasonable fit to the data.


### Check ETS diagnosis for ATM1

```{r}
checkresiduals(atm1_ets)
```

The residuals from the ETS(A,N,A) model fluctuate around zero with no obvious 
pattern, meaning we have a reasonable fit. The ACF plot shows that most correlations
lie within the confidence limits, so very little remaining autocorrelation. The 
Ljung-Box test gave a p-value of 0.2206, so we fail to reject the null hypothesis
of no autocorrelation. Once again, the residuals behave close to white noise, 
meaning the model is appropriate.

Since both models make sense, I will forecast with both ARIMA and ETS.


### Forecast ATM1 with both models

Forecast the next 31 days (end of May)

```{r}
atm1_arima_fc <- forecast(atm1_arima, h = 31)
atm1_ets_fc   <- forecast(atm1_ets, h = 31)

# plots
plot(atm1_arima_fc, main = "ATM1 ARIMA Forecast")
plot(atm1_ets_fc, main = "ATM1 ETS Forecast")
```

## ATM2 Forecast

### Fit ARIMA model for ATM2


```{r}
atm2_arima <- auto.arima(atm_ts$ATM2, seasonal = TRUE)
summary(atm2_arima)
```
The model select by `auto.arima()` for ATM2 is ARIMA(2,0,2)(0,1,2)[7] 


### Check ARIMA diagnosis for ATM2

```{r}
checkresiduals(atm2_arima)
```

The residuals from the ARIMA(2,0,2)(0,1,2)[7] model fluctuate around zero with 
no obvious pattern, so that means that the model has captured the main structure
of the data. The ACF plot shows that most correlations are within the confidence
limits, so barely any autocorrelation left. The Ljung-Box test produced a p-value 
of 0.1459, so we fail to reject the null hypothesis of no autocorrelation. The 
residuals behave like white noise, telling us the model is a good fit.

### Fit ETS for ATM2

```{r}
atm2_ets <- ets(atm_ts$ATM2)
summary(atm2_ets)
```

### Check ETS diagnosis for ATM2

```{r}
checkresiduals(atm2_ets)
```
We can see that residuals fluctuate around zero and most without a pattern, so
decent fit. The ACF has 3 correlations outside the limits, however, most correlations
are within limits. The Ljung-Box test gave a p-value of 0.0004379, so we fail to
reject the null hypothesis of no autocorrelation.


### Forecast ATM2 with both models

```{r}
atm2_arima_fc <- forecast(atm2_arima, h = 31)
atm2_ets_fc   <- forecast(atm2_ets, h = 31)

plot(atm2_arima_fc, main = "ATM2 ARIMA Forecast")
plot(atm2_ets_fc, main = "ATM2 ETS Forecast")
```
## Accuracy Testing
Since both models seem to be appropriate for both ATM1 and ATM2, I will compare
the accuracy on a test set using the 7 last observations as my test data.

### Split data into train and test set

ATM1

```{r}
atm1_train <- head(atm_ts$ATM1, length(atm_ts$ATM1) - 7)
atm1_test  <- tail(atm_ts$ATM1, 7)
```

ATM2

```{r}
atm2_train <- head(atm_ts$ATM2, length(atm_ts$ATM2) - 7)
atm2_test  <- tail(atm_ts$ATM2, 7)

```


### Fit models again on training data

ATM1

```{r}
atm1_arima_train <- auto.arima(atm1_train, seasonal = TRUE)
atm1_ets_train   <- ets(atm1_train)

atm1_arima_pred <- forecast(atm1_arima_train, h = 31)
atm1_ets_pred   <- forecast(atm1_ets_train, h = 31)
```

ATM2

```{r}
atm2_arima_train <- auto.arima(atm2_train, seasonal = TRUE)
atm2_ets_train   <- ets(atm2_train)

atm2_arima_pred <- forecast(atm2_arima_train, h = 31)
atm2_ets_pred   <- forecast(atm2_ets_train, h = 31)
```


### Forecast 

ATM1

```{r}
accuracy(atm1_arima_pred, atm1_test)
accuracy(atm1_ets_pred, atm1_test)
```

ATM2

```{r}
accuracy(atm2_arima_pred, atm2_test)
accuracy(atm2_ets_pred, atm2_test)
```

By analyzing the accuracy measures, we can see that for ATM1, both models 
performed similarly, but there are some differences since ARIMA had a slightly 
lower RMSE (6.51 vs 6.54) and ETS had slightly better MAE and MAPE. The performance
is very close, but I think ETS provides a marginally better fit because of its 
slightly lower error in MAE and MAPE.

For ATM2, it's a little different since ETS outperforms ARIMA across all key metrics.
The RMSE is lower (17.34 vs 20.59), the MAE is lower (12.93 vs 15.32), and 
MAPE is significantly lower (20.44 vs 98.63). So, for ATM2, the ETS is better suited,
probably because it handles variability and noise more effectively than ARIMA in
this case.

So, I selected ETS for both, although both models performed similarly for ATM1.


## ATM3 Forecast

As explained above, ATM3 is kind of a unique case, as the time series is mainly 
zeros with only a few observations by the end of the period. This probably means
that the ATM was inactive for most of the time frame. Because of this, standard 
time series models like ARIMA or ETS don't make sense since they rely on consistent
patterns to generate forecasts. I will instead use an average-based forecast.
I could have also excluded ATM3 because of the lack of data but I decided to do
it in case a full ATM analysis is preferred. 


```{r}
atm3_mean <- meanf(atm_ts$ATM3, h = 31)
plot(atm3_mean, main = "ATM3 Mean Forecast")
```

The forecast is pretty close to zero since the great majority of the values are
zero.

## ATM4 Forecast


As explained above, ATM4 has high variability and extreme spikes, telling us we 
have outliers and unstable variance, which can negatively impact standard time 
series models such as ARIMA and ETS. So, in this case, I will perform a log 
transformation to stabilize the variance and reduce the influence of outliers, to
allow the models to properly capture the actual structure of the data. After the log
transformation, I will use ARIMA and ETS and then use exponential to rever to the
original scale.


### Log Transform ATM4

```{r}
atm4_log <- log(atm_ts$ATM4 + 1)
plot(atm4_log, main = "ATM4 Log-Transformed")
```
The log transformation clearly reduced the impact of extremes (spikes).


### Fit ARIMA and ETS models in transformed data

ARIMA

```{r}
atm4_arima <- auto.arima(atm4_log, seasonal = TRUE)
summary(atm4_arima)
# residuals
checkresiduals(atm4_arima)
```

ETS

```{r}
atm4_ets <- ets(atm4_log)
summary(atm4_ets)

#residuals
checkresiduals(atm4_ets)
```


### Forecast

```{r}
atm4_arima_fc <- forecast(atm4_arima, h = 31)
atm4_ets_fc   <- forecast(atm4_ets, h = 31)
```


Convert back using exponential function

```{r}
atm4_arima_fc$mean  <- exp(atm4_arima_fc$mean) - 1
atm4_arima_fc$lower <- exp(atm4_arima_fc$lower) - 1
atm4_arima_fc$upper <- exp(atm4_arima_fc$upper) - 1
atm4_ets_fc$mean   <- exp(atm4_ets_fc$mean) - 1
atm4_ets_fc$lower <- exp(atm4_ets_fc$lower) - 1
atm4_ets_fc$upper <- exp(atm4_ets_fc$upper) - 1
```


### Plot

```{r}
plot(atm4_arima_fc, main = "ATM4 ARIMA Forecast")
plot(atm4_ets_fc, main = "ATM4 ETS Forecast")
```

I tried to forecast using manually transformed data (log transformation) but it 
produced inconsistent visualizations due to scaling issues. Instead, a log transformation
was applied directly within the functions, to make sure the forecasts could be 
properly interpreted in the original scale.

### Fit Models
 
```{r}
atm4_arima <- auto.arima(atm_ts$ATM4, lambda = 0)
atm4_ets   <- ets(atm_ts$ATM4, lambda = 0)
```

### Forecast

```{r}
atm4_arima_fc <- forecast(atm4_arima, h = 31)
atm4_ets_fc   <- forecast(atm4_ets, h = 31)
```

### Plots

```{r}
plot(atm4_arima_fc, main = "ATM4 ARIMA Forecast")
plot(atm4_ets_fc, main = "ATM4 ETS Forecast")

```
The plots for both ARIMA and ETS have very wide uncertainty intervals.


## Accuracy Test ATM4

I will perform accuracy outputs tests just like I did for ATM1 and ATM2.

### Split train and test sets


```{r}
atm4_train <- head(atm_ts$ATM4, length(atm_ts$ATM4) - 7)
atm4_test  <- tail(atm_ts$ATM4, 7)
```


### Fit models again on training data

```{r}
atm4_arima_train <- auto.arima(atm4_train, lambda = 0)
atm4_ets_train   <- ets(atm4_train, lambda = 0)
```


### Forecast

```{r}
atm4_arima_pred <- forecast(atm4_arima_train, h = 31)
atm4_ets_pred   <- forecast(atm4_ets_train, h = 31)
```


### Accuracy

```{r}
accuracy(atm4_arima_pred, atm4_test)
accuracy(atm4_ets_pred, atm4_test)
```

For ATM4, the forecast accuracy is way worse compared to the other ATMs, which was
expected because of the high variability in the data. Both ARIMA and ETS models 
produce considerably large errors, with very high RMSE and MAPE values.

Between the two models, ETS performs a little better, with lower RMSE (203 vs 213)
and MAE (155 vs 184), so it has more accurate predictions overall.

It is important to note that a major factor impacting performance is the extreme
spike in the data, which significantly increases variability and makes it difficult
for both models to produce stable forecasts. This explaines the higher uncertainty
and lower predictive accuracy.


### Final Forecasts Table

```{r}
atm1_final <- forecast(atm1_ets, h = 31)
atm2_final <- forecast(atm2_ets, h = 31)
atm3_final <- meanf(atm_ts$ATM3, h = 31)
atm4_final <- forecast(atm4_ets, h = 31)

may_dates <- seq(as.Date("2010-05-01"), as.Date("2010-05-31"), by = "day")

forecast_df <- data.frame(
  DATE = may_dates,
  ATM1 = as.numeric(atm1_final$mean),
  ATM2 = as.numeric(atm2_final$mean),
  ATM3 = as.numeric(atm3_final$mean),
  ATM4 = as.numeric(atm4_final$mean)
)
```

Export to Excel

```{r}
write_xlsx(forecast_df, "ATM_May_2010_Forecast.xlsx")
```




## Part A Conclusion

Since the four ATM datasets have different behaviors, I evaluated each individually.

For ATM1 and ATM2, both ARIMA and ETS models performed pretty well, capturing 
the underlying structure and weekly seasonality in the data. The performance for
ARIMA was good but ETS provided a slightly better fit overall and outperformed 
ARIMA for ATM2 based on forecast accuracy.

ATM3 contained mostly zero values with small activity toward the end of the time
period. Due to the lack of meaningful (or any) historical patterns, I used a simple 
average-based forecast.

ATM4 has high variability and extreme outliers, making forecasting more difficult.
I applied a log transformation to stabilize the variance, but both ARIMA and ETS
models have reduced accuracy. Among the two, ETS performed slightly better, although
results need to be interpreted with caution.


# Part B: Forecasting Power

## Load Data

```{r}
power_raw <- "https://raw.githubusercontent.com/JDO-MSDS/Data-624/refs/heads/main/Project%201/data/ResidentialCustomerForecastLoad-624%20-%20ResidentialCustomerForecastLoad.csv"

power <- read.csv(power_raw)

str(power)
head(power)
```

The dataset consists of monthly residential power consumption from January 1998 
to December 2013. I will convert the data into a time series object with a 
frequency of 12 to reflect monthly observations and capture yearly seasonality.

### Convert to time series

```{r}
power_ts <- ts(power$KWH, start = c(1998, 1), frequency = 12)

# plot
plot(power_ts, main = "Monthly Power Consumption", ylab = "KWH")

# clean
power_ts_clean <- na.interp(power_ts)

power_ts <- power_ts_clean
```

The plot shows a repeating seasonal pattern, with yearly clear peaks and dips. So,
we can say that residential electricity use is highly seasonal. Most of the series
stays within a fairly consistent range, despite some increase in variability towards
the end of the time period. There is a sharp drop right after 2010.


### Decomposition

```{r}
power_decomp <- decompose(power_ts)
plot(power_decomp)
```

I used decomposition to separate the series into trend, seasonal, and irregular 
components. The seasonal component has a consistent yearly pattern, confirming 
strong seasonality. The trend component indicates gradual changes in power 
consumption over time.


### ACF and PACF

```{r}
par(mfrow = c(1,2))

acf(power_ts, main = "ACF - Power")
pacf(power_ts, main = "PACF - Power")
```

The ACF plot shows strong and repeating spikes at regular intervals, particularly
around lag 1.0 years (12 months), so obvious yearly seasonality in the data. The
slow decay in autocorrelation suggests the presence of a trend.

The PACF plot has significant spikes at the first few lags, then a gradual decline,
indicating that autoregressive terms probably be useful in modeling the series.


### Train and test split sets

```{r}
train <- window(power_ts, end = c(2012, 12))
test  <- window(power_ts, start = c(2013, 1))
```


### Models

**ARIMA**

```{r}
power_arima <- auto.arima(train, seasonal = TRUE)
summary(power_arima)
checkresiduals(power_arima)
```

`auto.arima()` selected ARIMA(1,0,1)(2,1,0)[12] model with drift. The inclusion 
of seasonal terms reflects the strong yearly seasonality mentioned above.

The residual diagnostics indicate a good model fit. The Ljung-Box test produced 
a p-value of 0.9652, so no significant autocorrelation remains in the 
residuals. The model successfully captures the temporal structure of the data.

**ETS**

```{r}
power_ets <- ets(train)
summary(power_ets)
checkresiduals(power_ets)
```
The selected ETS model is ETS(M,N,M), which captures multiplicative seasonality 
without a trend. The model reflects the strong seasonal variation mentioned above.

Residual diagnostics show that the model gives an adequate fit, with a Ljung-Box
test p-value of 0.305. This means that the residuals behave close to white noise,
despite the fit is less strong compared to the ARIMA model.


### Forecast

```{r}
arima_fc <- forecast(power_arima, h = 12)
ets_fc   <- forecast(power_ets, h = 12)
```


### Accuracy

```{r}
accuracy(arima_fc, test)
accuracy(ets_fc, test)
```


I compared the forecasts using a 12 month test set (2013) to evaluate which one
is the best model. Both models capture the general seasonal pattern, but there are
differences in accuracy. The ETS model outperforms ARIMA across all key metrics 
on the test set since RMSE, MAE, and MAPE are lower.

So, despite the ARIMA model showed great residual diagnostics, the forecast accuracy
is noticeably worse than ETS. So, ETS is a better choice to capture the multiplicative
seasonal structure of the data.

Therefore, I selected ETS(M,N,M).


### 2014 Forecast

```{r}
final_forecast <- forecast(power_ets, h = 12)
plot(final_forecast)

# export to excel
forecast_2014 <- data.frame(
  Month = seq(as.Date("2014-01-01"), as.Date("2014-12-01"), by = "month"),
  Forecast = as.numeric(final_forecast$mean)
)

write_xlsx(forecast_2014, "Power_2014_Forecast.xlsx")
```

## Part B Conclucsion

The power consumption data set has strong yearly seasonality, which was effectively
captured by both ARIMA and ETS models. The ARIMA provided a good statistical fit
based on the residual diagnostics, however, ETS ended up having a superior forecasting
performance on the test set. Because of this, I selected ETS as the final model for predicting power consumption in 2014.


# Part C: Waterflow 

## Load Data

```{r}
pipe1_raw <- "https://raw.githubusercontent.com/JDO-MSDS/Data-624/refs/heads/main/Project%201/data/Waterflow_Pipe1%20-%20Waterflow_Pipe1.csv"
pipe1 <- read.csv(pipe1_raw)


pipe2_raw <- "https://raw.githubusercontent.com/JDO-MSDS/Data-624/refs/heads/main/Project%201/data/Waterflow_Pipe2%20-%20Waterflow_Pipe2.csv"
pipe2 <- read.csv(pipe2_raw)

str(pipe1)
head(pipe1)

str(pipe2)
head(pipe2)
```


## Convert datetime


```{r}
pipe1 <- pipe1 |>
  mutate(
    Date.Time = as.POSIXct(Date.Time * 86400, origin = "1899-12-30", tz = "UTC"),
    hour = as.POSIXct(format(Date.Time, "%Y-%m-%d %H:00:00"), tz = "UTC")
  )

pipe2 <- pipe2 |>
  mutate(
    Date.Time = as.POSIXct(Date.Time * 86400, origin = "1899-12-30", tz = "UTC"),
    hour = as.POSIXct(format(Date.Time, "%Y-%m-%d %H:00:00"), tz = "UTC")
  )
```


## Aggregate

Using the mean as suggested 

```{r}
pipe1_hourly <- pipe1 |>
  group_by(hour) |>
  summarise(WaterFlow = mean(WaterFlow, na.rm = TRUE))

pipe2_hourly <- pipe2 |>
  group_by(hour) |>
  summarise(WaterFlow = mean(WaterFlow, na.rm = TRUE))
```


## Plot

```{r}
plot(pipe1_hourly$hour, pipe1_hourly$WaterFlow,
     type = "l", main = "Pipe 1 Hourly Flow")

plot(pipe2_hourly$hour, pipe2_hourly$WaterFlow,
     type = "l", main = "Pipe 2 Hourly Flow")
```


The water flow data sets have timestamped observations recorded at irregular 
intervals. I converted the timestamp from Excel numeric format to standard datetime
format. To align the data, I aggregated the observations hourly. For hours with 
various readings, I used the average water flow. 

## Convert to a time series with 24 hours

```{r}
pipe1_ts <- ts(pipe1_hourly$WaterFlow, frequency = 24)
pipe2_ts <- ts(pipe2_hourly$WaterFlow, frequency = 24)
```


## Plot and ACF for both pipes time series

```{r}
plot(pipe1_ts, main = "Pipe 1 Time Series")
acf(pipe1_ts)

plot(pipe2_ts, main = "Pipe 2 Time Series")
acf(pipe2_ts)
```

The hourly series for both pipes 1 and 2 seem to fluctuate around a relatively 
constant level with no obvious long-term trend. The variability is fairly stable
over time. The ACF plots show that most autocorrelation values are within the 
confidence bounds after the first lag. So, both series behave close to white noise.
They are kind of stationary but they do not have strong patterns/structure, so
forecasting doesn't make much sense. 