---
title: "DATA 624 Homework 5"
author: "Joao De Oliveira"
date: "2026-03-10"
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readxl)
library(forecast)
library(ggplot2)
library(fpp2)
```

## Introduction
In this report, I will be answering questions 1, 5, 6, 7, and 8 in the 8th chapter
from the Hyndman book.


## Question 1
Figure 8.31 shows the ACFs for 36 random numbers, 360 random numbers and 1,000 
random numbers.

(a) Explain the differences among these figures. Do they all indicate that the 
data are white noise?

The three ACF plots look different mainly because of the different sample sizes.
The series with 36 observations looks much noisier, with larger mins and maxs, 
since small samples produce more random variation in the sample autocorrelations.
As the sample size increases to 360 and then 1,000 observations, the ACF values 
stay closer to zero and the plots look more stable. Even though the plots look 
different, all three are consistent with white noise since none of them show a 
clear pattern (increasing, decreasing, or seasonal structure). In fact, the 
autocorrelations fluctuate randomly around zero, which is what is expected 
from a white noise process.

(b) Why are the critical values at different distances from the mean of zero? 
Why are the autocorrelations different in each figure when they each refer to 
white noise?

The critical values are at different distances from zero because they depend on 
the sample size. The approximate critical values for white noise are + or - 1.96/√T,
where T is the number of observations. When the sample size is small, there is 
more sampling variability. As the sample size increases, the bounds become narrower
because the  estimates become more precise. Each plot comes from a different random
sample, which makes the white noise representations to be different. In theory, 
white noise has zero autocorrelation at all lags, but in practice the estimated 
autocorrelations will vary slightly due to random sampling. Larger samples produce
estimates that are closer to zero.


## Question 5
For your retail data (from Exercise 3 in Section 2.10), find the appropriate order
of differencing (after transformation if necessary) to obtain stationary data.


```{r}
retaildata_raw <- "https://raw.githubusercontent.com/JDO-MSDS/Data-624/refs/heads/main/Homework%205/retail%20-%20Data1.csv"
retaildata <- read.csv(retaildata_raw, skip = 1)
```


```{r}
myts <- ts(retaildata[,"A3349399C"],
           frequency = 12,
           start = c(1982,4))
```


```{r}
autoplot(myts) +
  labs(title = "Monthly Australian Clothing Retail Sales",
       y = "Retail Turnover")
```

Trasformation needed?

```{r}
lambda <- BoxCox.lambda(myts)
lambda
```

Since lambda is 0.02074707, which is very close to zero, I will apply a log 
transformation.


```{r}
myts_trans <- log(myts)
autoplot(myts_trans) +
  labs(title = "Log Transformed Clothing Retail Series",
       y = "log(Clothing Retail Turnover)")
```


The original retail series shows an upward trend and strong seasonal patterns.
The seasonal extremes become larger as the level of the series increases, indicating
that the variance grows over time. Because of this, I considered a Box-Cox 
transformation. The estimated parameter was 0.0207, which is very close to zero.
This suggests that a log transformation is appropriate.

After applying the log transformation, the seasonal fluctuations seem a little more
stable and the variance is more consistent over time. However, the transformed 
series still shows the same general trend and seasonality, but the scale is more
suitable for further time series analysis.

```{r}
ndiffs(myts_trans)
nsdiffs(myts_trans)
```

Since ndiffs = 1 and nsdiffs = 1, we need one seasonal difference (lag 12) and 
one regular difference. 

```{r}
myts_diff <- diff(myts_trans, lag = 12)
myts_diff2 <- diff(myts_diff)

autoplot(myts_diff2) +
   labs(title = "Differenced Log-Transformed Series",
       x = "Time",
       y = "Differenced log(Turnover)")
```

ACF of differenced series

```{r}
ggAcf(myts_diff2) +
  labs(title = "ACF of Differenced Log-Transformed Series")
```

The transformed series still shows both trend and seasonality, so we need difference
to make it stationary. Since the data is monthly but with a strong yearly seasonal
pattern, one seasonal difference with lag 12 makes sense to remove the upward trend.

After applying both one seasonal and one regular difference, the series fluctuates 
around a roughly constant mean. Also, now there is no strong seasonal pattern anymore. 
So, the appropriate differencing order is one regular difference and one 
seasonal difference at lag 12.


## Question 6
Use R to simulate and plot some data from simple ARIMA models.

(a) Use the following R code to generate data from an AR(1) model with  
ϕ1=0.6 and σ2=1. The process starts with y1=0.

```{r}
set.seed(1234)

y <- ts(numeric(100))
e <- rnorm(100)

for(i in 2:100)
  y[i] <- 0.6*y[i-1] + e[i]

autoplot(y) +
  labs(
    title = "Simulated AR(1) Series",
    x = "Time",
    y = "y"
  )
```
The simulated series shows persistence over time, meaning that observations tend
to be related to their recent past values. Because ϕ_{1}=0.6, the series tends to
move smoothly and values stay above or below zero for short periods before changing
direction.

(b) Produce a time plot for the series. How does the plot change as you change ϕ_{1}?


```{r}
simulate_ar1 <- function(phi, n = 100){
  y <- ts(numeric(n))
  e <- rnorm(n)
  
  for(i in 2:n)
    y[i] <- phi*y[i-1] + e[i]
  
  y
}

y1 <- simulate_ar1(0.2)
y2 <- simulate_ar1(0.5)
y3 <- simulate_ar1(0.8)

autoplot(cbind(phi_0.2 = y1,
               phi_0.5 = y2,
               phi_0.8 = y3),
         facets = TRUE) +
  labs(title = "AR1 Processes with Different Values of phi",
       x = "Time",
       y = "y")
```

By looking at the output plots for different ϕ, we can see that when ϕ~1~ is small (0.2),
the series seems like white noise, it looks random since it is constantly changing
direction. As ϕ~1~ increases, the series is smoother and with less changing in direction, meaning 
that values tend to remain above or below the mean for longer periods. For ϕ~1~=0.8, 
the series has long swings (trends) and a strong dependence on past values.


(c) Write your own code to generate data from an MA(1) model with $θ_{1}=0.6$ and 
$σ^2=1$.

MA(1) model: $y_{t}= ε_{t} + θ_{1}ε_{t-1}$ where $θ_{1}=0.6$ and $σ^2=1$


```{r}
set.seed(1234)

y_ma1 <- ts(numeric(100))
e <- rnorm(100)

for(i in 2:100){
  y_ma1[i] <- e[i] + 0.6 * e[i-1]
}

head(y_ma1, 10)
```


(d) Produce a time plot for the series. How does the plot change as you change θ~1~?


```{r}
autoplot(y_ma1) +
  labs(title = "Simulated MA(1) Series (theta = 0.6)",
       x = "Time",
       y = "y")
```

```{r}
simulate_ma1 <- function(theta, n = 100){
  y <- ts(numeric(n))
  e <- rnorm(n)
  
  for(i in 2:n){
    y[i] <- e[i] + theta * e[i-1]
  }
  
  y
}

set.seed(1234)

y1 <- simulate_ma1(0.2)
y2 <- simulate_ma1(0.5)
y3 <- simulate_ma1(0.8)

autoplot(cbind(theta_0.2 = y1,
               theta_0.5 = y2,
               theta_0.8 = y3),
         facets = TRUE) +
  labs(title = "MA(1) Processes with Different Values of theta",
       x = "Time",
       y = "y")
```

When θ~1~ is small, the series behaves similarly to white noise with little a
clear dependence between observations. As θ~1~ increases, neighboring observations
are more related to each other (short memory). However, unlike in the AR(1) process, the
effect of extremes does not persist for long periods, that's why the series still changes
direction frequently.


(e) Generate data from an ARMA(1,1) model with $ϕ_{1}=0.6$, $θ_{1}=0.6$ and 
$σ_{2}=1$.

ARMA(1,1) model: $y_{t} = ϕ_{1}y_{t-1} + ε_{t} + θ_{1}ε_{t-1}$


```{r}
set.seed(1234)

y_arma <- ts(numeric(100))
e <- rnorm(100)

for(i in 2:100){
  y_arma[i] <- 0.6*y_arma[i-1] + e[i] + 0.6*e[i-1]
}

# time plot
autoplot(y_arma) +
  labs(title = "Simulated ARMA(1,1) Series (phi = 0.6, theta = 0.6)",
       x = "Time",
       y = "y")
```

The ARMA(1,1) series combines characteristics of both AR and MA processes. The 
autoregressive component makes the series more persistence, while the MA component
adds short-term effects from recent shocks/behavior. So, the series seems smoother
than white noise but still has clear fluctuations over time.

(f) Generate data from an AR(2) model with $ϕ_{1}=−0.8$, $ϕ_{2}=0.3$ and $σ_{2}=1$.
(Note that these parameters will give a non-stationary series.)

AR(2) model: $y_{t} = ϕ_{1}y_{t-1} + ϕ_{2}y_{t-2} + ε_{t}$ 
with $ϕ_{1}=-0.8$ , $ϕ_{2}=0.3$ and $σ_{2}=1$ 

```{r}
set.seed(1234)

y_ar2 <- ts(numeric(100))
e <- rnorm(100)

for(i in 3:100){
  y_ar2[i] <- -0.8*y_ar2[i-1] + 0.3*y_ar2[i-2] + e[i]
}

# time plot
autoplot(y_ar2) +
  labs(title = "Simulated AR(2) Series (phi1 = -0.8, phi2 = 0.3)",
       x = "Time",
       y = "y")
```

The simulated AR(2) series has an oscillating pattern where the values repeatedly
switch between positive and negative (with the same amplitude). Over time, the 
oscillations become larger (more amplitude), causing the series to diverge away 
from zero, indicating that the process is non-stationary.


(g) Graph the latter two series and compare them.

```{r}
autoplot(cbind(ARMA11 = y_arma, AR2 = y_ar2), facets = TRUE) +
  labs(title = "Comparison of Simulated ARMA(1,1) and AR(2) Series",
       x = "Time",
       y = "y")
```
The two series have pretty difference behavior. The ARMA(1,1) series fluctuates 
around zero and remains relatively stable over time. The size of the fluctuations
is pretty consistent, which is typical of a stationary time series.

In contrast, the AR(2) series shows oscillations that grow larger over time. The
values constantly change between positive and negative, but the size of these swings
increases with time. This causes the series to diverge away from zero, creating
the growing behavior seen in the plot.


## Question 7
Consider wmurders, the number of women murdered each year (per 100,000 standard 
population) in the United States.

(a) By studying appropriate graphs of the series in R, find an appropriate 
ARIMA(p,d,q) model for these data.

```{r}
data(wmurders)
```

```{r}
autoplot(wmurders) +
  labs(title = "Women Murdered per 100,000 Population in the US",
       subtitle = "Original Time Series",
       x = "Year",
       y = "Murders per 100,000") +
  theme_minimal() +
  geom_point(color = "red", size = 2) +
  geom_line(color = "red", linewidth = 1)
```

The time series does not look stationary since the mean changes over time. The 
rate of women murdered rises from mid 1950s into the mid 1970s, stays relatively
high through the 1980s and early 1990s, and then declines noticeably toward the 
end of the series. Since the series behavior changes over time rather than 
fluctuating around a constant mean, differencing is needed before fitting an ARIMA
model.

```{r}
# first difference
wmurders_diff <- diff(wmurders)

autoplot(wmurders_diff) +
  labs(title = "First Differenced Women Murder Series",
       x = "Year",
       y = "Differenced rate")
```

The first differenced series fluctuates around a more stable mean now making first
difference appropriate.


Checking **ACF and PACF**

```{r}
ggAcf(wmurders_diff) +
  labs(title = "ACF of First-Differenced Series")

ggPacf(wmurders_diff) +
  labs(title = "PACF of First-Differenced Series")
```

### (a)

The original series is not stationary because its mean changes over time. The 
changing level explained above suggests that differencing is required. After taking
the first difference, the series fluctuates around a roughly constant mean with 
no clear trend, so one difference is sufficient (even though the ndiffs(wmurders)
gave me 2).

The ACF of the differenced series shows a notable spike at lag 2 while the other
lags fall mostly within the confidence bounds. The PACF also shows a significant
spike at lag 2. This pattern suggests an autoregressive structure of order 2 after
differencing. Therefore, an appropriate model for the series is:

ARIMA(2,1,0)

Confirming that the model makes sense

```{r}
Arima(wmurders, order = c(2,1,0))
```

Checking if residuals behave like white noise

```{r}
checkresiduals(Arima(wmurders, order = c(2,1,0)))
```
It is confirmed that the residuals behave like white noise.


(b) Should you include a constant in the model? Explain.

Because the model includes one difference (d = 1), adding a constant would imply
a linear trend in the original series. The time plot of the data does not show a
linear trend over the entire period, so including a constant is not appropriate.
Therefore, the ARIMA(2,1,0) model should be fitted without a constant.

(c) Write this model in terms of the backshift operator.

Since p=2, d=1, and q=0

My ARIMA model is: $ϕ(B)(1-B)y_{t}=ε_{t}$

The AR(2) polynomial is $ϕ(B)= 1 - ϕ_{1}B - ϕ_{2}B^2$

finding ϕ~1~ and ϕ~2~

```{r}
Arima(wmurders, order = c(2,1,0))
```

Since $ϕ_{1}=-0.0572$ and $ϕ_{2}=0.2967$

Then, $ϕ(B) = 1 - (-0.0572)B - 0.2967B^2$

If we substitute ϕ(B) in the ARIMA model:
$(1 + 0.0572B - 0.2967B^2)(1-B)y_{t}=ε_{t}$


(d) Fit the model using R and examine the residuals. Is the model satisfactory?

Fit the model (as I did above)

```{r}
fit_wmurders <- Arima(wmurders, order = c(2,1,0))
fit_wmurders
```

Check Residuals

```{r}
checkresiduals(fit_wmurders)
```


The residual time plot shows that the residuals fluctuate randomly around zero 
with no obvious trend or pattern. The residual ACF shows that autocorrelations lie
within the significance bounds. Also, the histogram of the residuals is approximately
symmetric and bell-shaped, suggesting that the residuals are roughly normally distributed.

Overall, these results suggest that the residuals behave like white noise, 
indicating that the ARIMA(2,1,0) model gives a satisfactory fit to the data.


(e) Forecast three times ahead. Check your forecasts by hand to make sure that you
know how they have been calculated.


```{r}
fc_wmurders <- forecast(fit_wmurders, h = 3)
fc_wmurders
autoplot(fc_wmurders) +
  labs(title = "Three-Step-Ahead Forecasts for Women Murder Rate",
       x = "Year",
       y = "Murder rate")
```


Checking by hand, which means to 

```{r}
# last observed values
yT   <- tail(wmurders, 1)
yT1  <- tail(wmurders, 2)[1]
yT2  <- tail(wmurders, 3)[1]

# last two differences
wT   <- yT - yT1
wT1  <- yT1 - yT2

# AR coefficients 
phi1 <- coef(fit_wmurders)["ar1"]
phi2 <- coef(fit_wmurders)["ar2"]

# forecasts differenced series
w_hat1 <- phi1 * wT + phi2 * wT1
w_hat2 <- phi1 * w_hat1 + phi2 * wT
w_hat3 <- phi1 * w_hat2 + phi2 * w_hat1

# convert back to original scale
y_hat1 <- yT + w_hat1
y_hat2 <- y_hat1 + w_hat2
y_hat3 <- y_hat2 + w_hat3

c(y_hat1, y_hat2, y_hat3)
fc_wmurders$mean
```

Using the estimated coefficients ar1 and ar2, each forecast depends on the two 
most recent differenced values. The 1 step ahead forecast is calculated first, and 
the two-step and three-step forecasts are then obtained using the previous one 
using the earlier forecasted differences. Converting these back to the original 
scale gives the final forecasts for the murder rate.

The hand calculations match the values produced by R, which confirms that the 
forecasts are accurate.


(f) Create a plot of the series with forecasts and prediction intervals for the next three periods shown.

```{r}
fc_wmurders <- forecast(fit_wmurders, h = 3)

autoplot(fc_wmurders) +
  labs(title = "Women Murder Rate with 3-Step-Ahead Forecasts",
       x = "Year",
       y = "Murder rate")
```


(g) Does auto.arima() give the same model you have chosen? If not, which model do you think is better?


```{r}
auto.arima(wmurders)
```

The `auto.arima()` function selected an ARIMA(1,2,1) model, which is different than
the ARIMA(2,1,0) I selected. The automatic model uses two differences, which may
indicate over-differencing because the first-differenced series already seemed 
stationary. Since the ARIMA(2,1,0) model produces satisfactory residual diagnostics
and aligns with the behavior observed in the ACF and PACF plots, it is the preferred
model.


## Question 8
Consider austa, the total international visitors to Australia (in millions) for the period 1980-2015.

(a) Use auto.arima() to find an appropriate ARIMA model. What model was selected. Check that the residuals look like white noise. Plot forecasts for the next 10 periods.


```{r}
fit_austa <- auto.arima(austa)
fit_austa
```

The model selected by auto.arima() is ARIMA(0,1,1) model with drift. The MA(1) 
coefficient is 0.3006 and the drift term is 0.1735, meaning that there is a 
gradual upward trend in the number of international visitors to Australia over time.

Checking Residuals

```{r}
checkresiduals(fit_austa)
```
We can confirm that the residuals look like white noise and that there is no
significant autocorrelation remaining. So, the model properly captures the structure
of the series.


```{r}
forecast_austa <- forecast(fit_austa, h = 10)

autoplot(forecast_austa) +
  labs(title = "Forecasts of International Visitors to Australia",
       x = "Year",
       y = "Visitors [millions]")
```

Forecasts for the next 10 periods show an upward trend in the number of international
visitors to Australia. Naturally, the prediction intervals are larger with time (more uncertainty).


(b) Plot forecasts from an ARIMA(0,1,1) model with no drift and compare these to 
part a. Remove the MA term and plot again.


ARIMA(0,1,1) with no drift
```{r}
fit_austa_nodrift <- Arima(austa, order = c(0,1,1), include.drift = FALSE)
fit_austa_nodrift

# plot with forecast
fc_austa_nodrift <- forecast(fit_austa_nodrift, h = 10)

autoplot(fc_austa_nodrift) +
  labs(title = "ARIMA(0,1,1) Forecasts Without Drift",
       x = "Year",
       y = "Visitors [millions]")
```

Now the forecasts without drift are flat when compared with (a) because the 
model no longer includes a trend term. In part (a), the drift term produced forecasts
that continued to increase over time. Without drift, the forecasts are more level
and do not consider the long-run upward movement as strongly.

If we remove the MA: ARIMA(0,1,0)

```{r}
fit_austa_r <- Arima(austa, order = c(0,1,0), include.drift = FALSE)
fit_austa_r

# plot with forecasts
fc_austa_r <- forecast(fit_austa_r, h = 10)

autoplot(fc_austa_r) +
  labs(title = "ARIMA(0,1,0) Forecasts Without Drift",
       x = "Year",
       y = "Visitors [millions]")
```

The forecasts from the ARIMA(0,1,1) model without drift and the ARIMA(0,1,0) model
seem very similar. In both cases, the forecasts remain approximately flat at the
last observed value of the series. This happens because both models use one 
difference and do not include a drift term. Without drift, the models do not 
project a trend, so the best forecast for future values is simply the most recent
observation. The ARIMA(0,1,1) model accounts for short-term dependence through the
MA term, while the ARIMA(0,1,0) model is a simple random walk. 

(c) Plot forecasts from an ARIMA(2,1,3) model with drift. Remove the constant and see what happens.


```{r}
fit_austa_213_drift <- Arima(austa, order = c(2,1,3), include.drift = TRUE)
fit_austa_213_drift

# plot forecasts
fc_austa_213_drift <- forecast(fit_austa_213_drift, h = 10)

autoplot(fc_austa_213_drift) +
  labs(title = "ARIMA(2,1,3) Forecasts with Drift",
       x = "Year",
       y = "Visitors [millions]")
```

Without constant

```{r}
fit_austa_213_nodrift <- Arima(austa, order = c(2,1,3),
                               include.drift = FALSE,
                               method = "ML")
fit_austa_213_nodrift

# plot forecasts
fc_austa_213_nodrift <- forecast(fit_austa_213_nodrift, h = 10)

autoplot(fc_austa_213_nodrift) +
  labs(title = "ARIMA(2,1,3) Forecasts without Drift",
       x = "Year",
       y = "Visitors [millions]")
```

The forecasts for the ARIMA(2,1,3) model without drift still show an increasing 
trend even though the constant term was removed. This happens because the 
autoregressive structure, particularly the second AR coefficient which is close 
to 1, produces strong persistence in the series.


(d) Plot forecasts from an ARIMA(0,0,1) model with a constant. 
Remove the MA term and plot again.

ARIMA(0,0,1) with a constant

```{r}
fit_austa_001 <- Arima(austa, order = c(0,0,1), include.mean = TRUE)
fit_austa_001

# plot forecasts
fc_austa_001 <- forecast(fit_austa_001, h = 10)

autoplot(fc_austa_001) +
  labs(title = "ARIMA(0,0,1) Forecasts with Constant",
       x = "Year",
       y = "Visitors [millions]")
```

Remove MA: ARIMA(0,0,0)

```{r}
fit_austa_000 <- Arima(austa, order = c(0,0,0), include.mean = TRUE)
fit_austa_000

# plot forecasts
fc_austa_000 <- forecast(fit_austa_000, h = 10)

autoplot(fc_austa_000) +
  labs(title = "ARIMA(0,0,0) Forecasts with Constant",
       x = "Year",
       y = "Visitors [millions]")
```

The ARIMA(0,0,1) model with a constant produces forecasts that quickly gets close
to the estimated mean of about 3.55 million. The MA term allows a short adjustment
from the last observed value before the forecasts approaches the mean.

When the MA term is removed, the ARIMA(0,0,0) model simply forecasts the mean 
for all future periods. Both models ignore the strong upward trend in the data.


(e) Plot forecasts from an ARIMA(0,2,1) model with no constant.

ARIMA(0,2,1) with no constant

```{r}
fit_austa_021 <- Arima(austa, order = c(0,2,1), include.constant = FALSE)
fit_austa_021

# plot forecasts
fc_austa_021 <- forecast(fit_austa_021, h = 10)

autoplot(fc_austa_021) +
  labs(title = "ARIMA(0,2,1) Forecasts without Constant",
       x = "Year",
       y = "Visitors [millions]")
```
The ARIMA(0,2,1) model without a constant produces forecasts that continue the 
upward movement in the series. The prediction intervals widen rapidly, indicating
increasing uncertainty in the future, as expected.


## Question 9
For the usgdp series:

(a) if necessary, find a suitable Box-Cox transformation for the data;


```{r}
# plot to see behavior
autoplot(usgdp) +
  labs(title = "US GDP Time Series",
       x = "Year",
       y = "GDP")

# box cox parameter
lambda <- BoxCox.lambda(usgdp)
lambda
```


The plot of the usgdp series shows a strong upward trend and increasing variability
over time, suggesting that a transformation may be useful to stabilize the variance.

The Box-Cox transformation parameter was estimated using BoxCox.lambda(), which gave
lambda = 0.366. Since this value is between 0 and 1, a Box-Cox transformation is
appropriate to reduce the increasing variance in the series.

(b) fit a suitable ARIMA model to the transformed data using auto.arima();

```{r}
# using auto arima with previous calculated lambda
fit_usgdp <- auto.arima(usgdp, lambda = lambda)
fit_usgdp
```

The auto.arima() function selected ARIMA(2,1,0) with drift and got the following
coefficients:
Estimated coefficients:

- AR(1) = 0.2795  
- AR(2) = 0.1208  
- Drift = 0.1829 

This means that the GDP series required one difference (d = 1) to remove the trend,
and the remaining dynamics are captured by two autoregressive terms. The drift 
term represents the long-term upward growth in GDP.

(c) try some other plausible models by experimenting with the orders chosen;

Trying ARIMA(1,1,0), ARIMA(0,1,1), and ARIMA(1,1,1)

```{r}
fit1 <- Arima(usgdp, order = c(1,1,0), lambda = lambda)
fit2 <- Arima(usgdp, order = c(0,1,1), lambda = lambda)
fit3 <- Arima(usgdp, order = c(1,1,1), lambda = lambda)

fit1
fit2
fit3
```

I tested ARIMA(1,1,0), ARIMA(0,1,1), and ARIMA(1,1,1).

The resulting AIC values were:

- ARIMA(1,1,0): AIC = −64.78  
- ARIMA(0,1,1): AIC = −10.79  
- ARIMA(1,1,1): AIC = −86.43  

ARIMA(1,1,1) provided the best fit of the tested models, as it had the lowest AIC.
However, the model selected earlier by auto.arima() (ARIMA(2,1,0) with drift - AIC = −115.11)
still provides a better fit overall because it has a substantially lower AIC.

So, the ARIMA(2,1,0) with drift still is the preferred model for the transformed 
GDP series.


(d) choose what you think is the best model and check the residual diagnostics;

As explained above, the best model is ARIMA(2,1,0) since it has the lowest AIC.

Residuals for ARIMA(2,1,0)

```{r}
checkresiduals(fit_usgdp)
```

The residual time plot shows that the residuals fluctuate randomly around zero 
with no obvious patterns. The histogram of residuals appears approximately normally
distributed. Also, the residual ACF shows no significant spikes outside the 
confidence bounds, which means that there is little remaining autocorrelation.

The Ljung–box test gives a p-value of 0.3617 (>0.05). Sp, we fail to reject the 
null hypothesis that the residuals are independent.

The residual diagnostics tells us that the ARIMA(2,1,0) with drift model provides
an adequate fit to the transformed GDP series.


(e) produce forecasts of your fitted model. Do the forecasts look reasonable?


```{r}
usgdp_fc <- forecast(fit_usgdp, h = 10)
autoplot(usgdp_fc)
```

The forecasts show a continuation of the upward trend in US GDP. The prediction 
intervals are larger as the forecast horizon increases, showing a greater 
uncertainty in longer-term forecasts, as expected. The forecasts appear reasonable
because they follow the long-term growth trend present in the historical GDP series 
and considering the greater uncertainty over time.

(f) compare the results with what you would obtain using ets() (with no transformation).

```{r}
fit_ets <- ets(usgdp)
fit_ets

usgdp_ets_fc <- forecast(fit_ets, h = 10)
autoplot(usgdp_ets_fc)
```


Using ets(usgdp), the selected model was ETS(A,A,N) (additive error, additive trend,
no seasonality). The ETS forecasts also show a continued upward trend in US GDP,
similar to the forecasts from the ARIMA(2,1,0) with drift model. Both models seem
tocapture the strong growth pattern in the data, and the forecasts appear very 
similar. The main difference is that the ARIMA model used a Box-Cox transformation
and differencing, while the ETS model directly models the level and trend components
of the series.