---
title: "DATA 624 Homework 5"
author: "Joao De Oliveira"
date: "2026-03-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Introduction
In this report, I will be answering questions 1, 5, 6, 7, and 8 in the 8th chapter
from the Hyndman book.


## Question 1
Figure 8.31 shows the ACFs for 36 random numbers, 360 random numbers and 1,000 
random numbers.

(a) Explain the differences among these figures. Do they all indicate that the 
data are white noise?

The three ACF plots look different mainly because of the different sample sizes.
The series with 36 observations looks much noisier, with larger mins and maxs, 
since small samples produce more random variation in the sample autocorrelations.
As the sample size increases to 360 and then 1,000 observations, the ACF values 
stay closer to zero and the plots look more stable. Even though the plots look 
different, all three are consistent with white noise since none of them show a 
clear pattern (increasing, decreasing, or seasonal structure). In fact, the 
autocorrelations fluctuate randomly around zero, which is what is expected 
from a white noise process.

(b) Why are the critical values at different distances from the mean of zero? 
Why are the autocorrelations different in each figure when they each refer to 
white noise?

The critical values are at different distances from zero because they depend on 
the sample size. The approximate critical values for white noise are + or - 1.96/√T,
where T is the number of observations. When the sample size is small, there is 
more sampling variability. As the sample size increases, the bounds become narrower
because the  estimates become more precise. Each plot comes from a different random
sample, which makes the white noise representations to be different. In theory, 
white noise has zero autocorrelation at all lags, but in practice the estimated 
autocorrelations will vary slightly due to random sampling. Larger samples produce
estimates that are closer to zero.


## Question 5
For your retail data (from Exercise 3 in Section 2.10), find the appropriate order
of differencing (after transformation if necessary) to obtain stationary data.


```{r}
library(readxl)
library(forecast)
library(ggplot2)

retaildata <- read_excel("C:\\Users\\jpfer\\Downloads\\retail.xlsx", skip = 1)
```


```{r}
myts <- ts(retaildata[,"A3349399C"],
           frequency = 12,
           start = c(1982,4))
```


```{r}
autoplot(myts) +
  labs(title = "Monthly Australian Clothing Retail Sales",
       y = "Retail Turnover")
```

Trasformation needed?

```{r}
lambda <- BoxCox.lambda(myts)
lambda
```

Since lambda is 0.02074707, which is very close to zero, I will apply a log 
transformation.


```{r}
myts_trans <- log(myts)
autoplot(myts_trans) +
  labs(title = "Log Transformed Clothing Retail Series",
       y = "log(Clothing Retail Turnover)")
```


The original retail series shows an upward trend and strong seasonal patterns.
In the original plot, the seasonal extremes become larger as the level of the 
series increases, indicating that the variance grows over time. Because of this, I
considered a Box-Cox transformation. The estimated parameter was 0.0207, which is
very close to zero. This suggests that a log transformation is appropriate.

After applying the log transformation, the seasonal fluctuations seem more stable
and the variance is more consistent over time. The transformed series still shows
the same general trend and seasonality, but the scale is more suitable for further 
time series analysis.

```{r}
ndiffs(myts_trans)
nsdiffs(myts_trans)
```

Since ndiffs = 1 and nsdiffs = 1, we need 1 seasonal difference (lag 12) and 
1 regular difference. 

```{r}
myts_diff <- diff(myts_trans, lag = 12)
myts_diff2 <- diff(myts_diff)

autoplot(myts_diff2)
```

