---
title: "DATA 624 Homework 6"
author: "Joao De Oliveira"
date: "2026-03-21"
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(fpp2)
```

## Introduction
In this report, I will be answering questions 1, 2, 3, 5, 6, 7, and 8 in the 9th chapter
from the Hyndman book.


## Question 1
Consider monthly sales and advertising data for an automotive parts company (data set advert).

(a) Plot the data using autoplot. Why is it useful to set facets=TRUE?

Load Data

```{r}
advert
```

```{r}
autoplot(advert, facets=TRUE)
```

It is helpful to set facets=TRUE because the dataset has 2 variables (advert and
sales) with different ranges, so they should have different scales. Using facets=TRUE
plots the variables in different panels with the appropriate scale for each variable.
It is also much easier to understand and compare patterns over time. 


(b) Fit a standard regression model  $y_{t}=a+b_{x_{t}}+η_{t}$ where yt denotes sales and xt 
denotes advertising using the tslm() function.

```{r}
fit_reg <- tslm(sales ~ advert)
summary(fit_reg)
```