---
title: "Data 624 Homework 2"
author: "Joao De Oliveira"
date: "2026-02-21"
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(fpp3)
```

## Introduction
In this report, I will be answering questions 1, 2, 3, 4, and 7 from the
5.11 section from the Hyndman online Forecasting book


## Question 1
Produce forecasts for the following series using whichever of NAIVE(y), SNAIVE(y) or RW(y ~ drift()) is more appropriate in each cas-e:

- Australian Population (global_economy)
- Bricks (aus_production)
- NSW Lambs (aus_livestock)
- Household wealth (hh_budget).
- Australian takeaway food turnover (aus_retail).


```{r exercise1-1}
aus_pop <- global_economy |>
  filter(Country == "Australia") |>
  select(Year, Population)

aus_pop |>
  model(RWdrift = RW(Population ~ drift())) |>
  forecast(h = 10) |>
  autoplot(aus_pop)
```


```{r exercise1-2}
bricks_clean <- bricks |>
  filter(!is.na(Bricks))

fit_bricks <- bricks_clean |>
  model(SNaive = SNAIVE(Bricks))

fc_bricks <- fit_bricks |>
  forecast(h = "2 years")

autoplot(fc_bricks, bricks_clean) +
  labs(title = "Bricks — Seasonal Naive Forecast (2 years)",
       x = "Quarter", y = "Bricks")
```


```{r exercise1-3}
nsw_lambs <- aus_livestock |>
  filter(State == "New South Wales", Animal == "Lambs") |>
  select(Month, Count)

nsw_lambs |>
  model(SNaive = SNAIVE(Count)) |>
  forecast(h = "2 years") |>
  autoplot(nsw_lambs)
```


```{r exercise1-4}
hh_budget |>
  model(RWdrift = RW(Wealth ~ drift())) |>
  forecast(h = 10) |>
  autoplot(hh_budget)
```


```{r exercise1-5}
takeaway_aus <- aus_retail |>
  filter(Industry == "Cafes, restaurants and takeaway food services") |>
  index_by(Month) |>                         
  summarise(Turnover = sum(Turnover, na.rm = TRUE))

takeaway_aus |>
  model(SNaive = SNAIVE(Turnover)) |>
  forecast(h = "2 years") 
  autoplot(takeaway_aus, Turnover) +
  labs(title = "Australian Takeaway Food Turnover — Seasonal Naive Forecast")
```

In this case I aggregated all the states (summing all states' monthly turnover) 
to be able to plot entire Australia's turnover series. 


Australian population and household wealth exhibit clear long-term trends with no
seasonality, so I used random walk models with drift. Bricks production, NSW lamb
slaughter, and takeaway food turnover display strong seasonal patterns, making 
seasonal naïve forecasts more appropriate. These models provide simple baseline 
forecasts that preserve the dominant features of each series.


## Question 2
Use the Facebook stock price (data set gafa_stock) to do the following:

a. Produce a time plot of the series.

```{r exercise2-a}
facebook <- gafa_stock |>
  filter(Symbol == "FB")

autoplot(facebook, Close) +
  labs(title = "Facebook Stock Daily Closing Price",
       x = "Date", y = "Closing Price [USD]")
```

The Facebook stock price shows an overall upward trend with substantial day-to-day
volatility and no clear seasonality, which is typical for financial time series.

b. Produce forecasts using the drift method and plot them.

```{r exercise2-b}
# facebook trading-day data
fb <- gafa_stock |>
  filter(Symbol == "FB") |>
  select(Date, Close) |>
  arrange(Date) |>
  as_tibble()

# daily calendar and join prices onto it
fb_daily_tbl <- tibble(Date = seq(min(fb$Date), max(fb$Date), by = "day")) |>
  left_join(fb, by = "Date") |>
  tidyr::fill(Close, .direction = "down")   # carry last trading close forward

# regular tsibble
fb_daily <- fb_daily_tbl |>
  as_tsibble(index = Date) |>
  tsibble::update_tsibble(regular = TRUE)

fc_drift <- fb_daily |>
  model(RWdrift = RW(Close ~ drift())) |>
  forecast(h = 30)

autoplot(fc_drift, fb_daily) +
  labs(title = "Facebook Closing Price — Drift Forecast (30 days, daily regularized)",
       x = "Date", y = "Close (USD)")
```


c. Show that the forecasts are identical to extending the line drawn between the first and last observations.

```{r exercise2-c}
# compute slope between first and last obs
y1 <- fb_daily |> as_tibble() |> slice(1) |> pull(Close)
yT <- fb_daily |> as_tibble() |> slice(n()) |> pull(Close)
Tn <- nrow(fb_daily)
slope <- (yT - y1) / (Tn - 1)

fc_tbl <- fc_drift |>
  as_tibble() |>
  select(Date, drift_mean = .mean)

ggplot() +
  geom_line(data = fb_daily |> as_tibble(), aes(Date, Close)) +
  geom_line(data = fc_tbl, aes(Date, drift_mean)) +
  geom_abline(intercept = y1, slope = slope, linetype = "dashed") +
  labs(title = "Drift forecast equals extension of the first–last line",
       x = "Date", y = "Close (USD)")
```

d. Try using some of the other benchmark functions to forecast the same data set. Which do you think is best? Why?


