---
title: "Data 624 Homework 8"
author: "Joao De Oliveira"
date: "2026-05-01"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
In this report, I will be answering questions 2 and 5 in the 7th chapter
from the Kuhn and Johnson book.


## Libraries

```{r packages}
library(mlbench)
library(caret)
library(earth)       
library(kernlab)    
library(ggplot2)
library(tidyr)
library(gridExtra)
```


### Question 7.2
Friedman (1991) introduced several benchmark data sets create by simulation. One of these simulations used the following nonlinear equation to create data:
y = 10 sin(πx1x2) + 20(x3 − 0.5)2 + 10x4 + 5x5 + N(0, σ2)
where the x values are random variables uniformly distributed between [0, 1]
(there are also 5 other non-informative variables also created in the simulation). The package mlbench contains a function called mlbench.friedman1 that simulates these data:


```{r exercise7-2}
set.seed(200)

trainingData <- mlbench.friedman1(200, sd = 1)
trainingData$x <- data.frame(trainingData$x)

testData <- mlbench.friedman1(5000, sd = 1)
testData$x <- data.frame(testData$x)

ctrl <- trainControl(method = "boot", number = 25)
```

**KNN**

```{r knn}  
set.seed(200)
knnModel <- train(x = trainingData$x,
                  y = trainingData$y,
                  method = "knn",
                  preProc = c("center", "scale"),
                  tuneLength = 10,
                  trControl = ctrl)
knnModel
knnPred <- predict(knnModel, newdata = testData$x)
knn_res <- postResample(pred = knnPred, obs = testData$y)
cat("KNN Test Performance:\n"); print(round(knn_res, 4))
```

**MARS**

```{r mars}
set.seed(200)
marsGrid <- expand.grid(degree = 1:2, nprune = 2:20)
marsModel <- train(x = trainingData$x,
                   y = trainingData$y,
                   method = "earth",
                   tuneGrid = marsGrid,
                   trControl = ctrl)
marsModel
marsPred <- predict(marsModel, newdata = testData$x)
mars_res <- postResample(pred = marsPred, obs = testData$y)
cat("MARS Test Performance:\n"); print(round(mars_res, 4))

# does it select only X1-X5?
varImp(marsModel)
```

**SVM**

```{r svm}
set.seed(200)
svmModel <- train(x = trainingData$x,
                  y = trainingData$y,
                  method = "svmRadial",
                  preProc = c("center", "scale"),
                  tuneLength = 14,
                  trControl = ctrl)
svmModel
svmPred <- predict(svmModel, newdata = testData$x)
svm_res <- postResample(pred = svmPred, obs = testData$y)
cat("SVM Test Performance:\n"); print(round(svm_res, 4))
```

**Neural Network**

```{r neural-network}
set.seed(200)
nnetGrid <- expand.grid(size  = c(1, 3, 5, 7, 9),
                        decay = c(0, 0.001, 0.01, 0.1))
nnetModel <- train(x = trainingData$x,
                   y = trainingData$y,
                   method = "nnet",
                   preProc = c("center", "scale", "spatialSign"),
                   tuneGrid = nnetGrid,
                   trControl = ctrl,
                   linout  = TRUE,
                   trace   = FALSE,
                   MaxNWts = 10 * (ncol(trainingData$x) + 1) + 10 + 1,
                   maxit   = 500)
nnetModel
nnetPred <- predict(nnetModel, newdata = testData$x)
nnet_res <- postResample(pred = nnetPred, obs = testData$y)
cat("NNet Test Performance:\n"); print(round(nnet_res, 4))
```

**Summary Table**

```{r summary}
results_72 <- data.frame(
  Model     = c("KNN", "MARS", "SVM Radial", "Neural Net"),
  Test_RMSE = round(c(knn_res["RMSE"], mars_res["RMSE"],
                      svm_res["RMSE"], nnet_res["RMSE"]), 4),
  Test_Rsq  = round(c(knn_res["Rsquared"], mars_res["Rsquared"],
                      svm_res["Rsquared"], nnet_res["Rsquared"]), 4)
)
print(results_72)
```

**Plots**

```{r plots}
# plot: test RMSE comparison
ggplot(results_72, aes(x = reorder(Model, -Test_RMSE), y = Test_RMSE, fill = Model)) +
  geom_col(width = 0.6, color = "white") +
  geom_text(aes(label = round(Test_RMSE, 3)), vjust = -0.4, size = 4.5, fontface = "bold") +
  scale_fill_manual(values = c("KNN"       = "#e07b54",
                               "MARS"      = "#4CAF50",
                               "SVM Radial"= "#5b8dd9",
                               "Neural Net"= "#9c6fd6")) +
  labs(title = "Q7.2 – Test Set RMSE by Model (Friedman1 Data)",
       subtitle = "Lower RMSE = Better | MARS clearly dominates",
       x = "Model", y = "Test RMSE") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none", plot.title = element_text(face = "bold"))

# plot MARS variable importance
vi72 <- data.frame(
  Predictor   = c("X1","X4","X2","X5","X3","X6","X7","X8","X9","X10"),
  Importance  = c(100, 75.33, 48.88, 15.63, 0, 0, 0, 0, 0, 0),
  Informative = c(TRUE, TRUE, TRUE, TRUE, TRUE, FALSE, FALSE, FALSE, FALSE, FALSE)
)
ggplot(vi72, aes(x = reorder(Predictor, Importance), y = Importance, fill = Informative)) +
  geom_col(width = 0.6, color = "white") +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "#4CAF50", "FALSE" = "#e07b54"),
                    labels  = c("Non-informative (X6–X10)", "Informative (X1–X5)")) +
  labs(title    = "Q7.2 – MARS Variable Importance",
       subtitle = "MARS correctly selects only informative predictors X1–X5",
       x = "Predictor", y = "Relative Importance (%)", fill = "Variable Type") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```